---
title: "MC1R Progression Analysis"
author: "Jackson G. Schumacher"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
---

<!-- Produces: Table 1, Tables 2-3, Supplemental Tables 1-9 and 11, Figure 1 (PPMI motor
     decline), Figure 3 (phenoconversion), Supplemental Figure 1 (PC-adjusted ancestry
     sensitivity), and PPMI supplemental diagnostic figures (slope distribution,
     slope vs. time to treatment, attrition) -->
<!-- Manuscript: Schumacher et al., Common MC1R variants and Parkinson disease progression -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = normalizePath(".."))
set.seed(10719L)
```

```{r packages}
library(readxl)
library(readr)
library(dplyr)
library(tidyr)
library(nlme)
library(survival)
library(ggplot2)
library(openxlsx)
library(ggtext)
library(pammtools)
library(patchwork)
```

```{r config}
config <- list(
  data      = "Data/MC1R_Combined_Data_Filtered.xlsx",
  updrs2    = "Data/MDS_UPDRS_Part_II__Patient_Questionnaire_29Jan2026.csv",
  updrs3    = "Data/MDS-UPDRS_Part_III_26Feb2026.csv",
  scopa_aut = "Data/SCOPA-AUT_24Feb2026.csv",
  gds       = "Data/Geriatric_Depression_Scale__Short_Version__24Feb2026.csv",
  ess       = "Data/Epworth_Sleepiness_Scale_26Feb2026.csv",
  out_dir   = "Output",
  pcs       = "Data/ppmi_pca_eigenvec.csv"
)
if (!dir.exists(config$out_dir)) dir.create(config$out_dir)

SH <- list(
  demographics = "Demographics",     updrs3       = "UPDRS_III",
  updrs1       = "UPDRS_I",          updrs2       = "UPDRS_II",
  updrs4       = "UPDRS_IV",         moca         = "MoCA",
  dat          = "DAT_SPECT",        csf          = "CSF_Tau_pTau_Abeta",
  asyn         = "CSF_AlphaSyn",     saa_detailed = "SAA_Kinetics_Detailed",
  nfl_gfap     = "NfL_GFAP",        updrs3_on    = "UPDRS_III_ON",
  updrs3_off_sub = "UPDRS_III_OFF_Subscores",
  updrs3_on_sub  = "UPDRS_III_ON_Subscores"
)

VALID_COHORTS <- c("SPD", "MC1R PD", "LRRK2 PD", "GBA PD",
                   "MC1R + LRRK2 PD", "MC1R + GBA PD")

COMPARISONS <- list(
  primary        = list(test = "MC1R PD",          ref = "SPD"),
  lrrk2_modifier = list(test = "MC1R + LRRK2 PD", ref = "LRRK2 PD"),
  gba_modifier   = list(test = "MC1R + GBA PD",   ref = "GBA PD")
)

VARIANT_COLS <- c(
  "R163Q" = "Arg163Gln/Pro (rs885479)",
  "V60L"  = "Val60Leu (rs1805005)",
  "V92M"  = "Val92Met/Leu (rs2228479)",
  "R151C" = "Arg151Ser/Gly/Cys (rs1805007)",
  "R160W" = "Arg160Trp (rs1805008)",
  "D294H" = "Asp294Asn/His (rs1805009)",
  "D84E"  = "Asp84Glu (rs1805006)",
  "R142H" = "Arg142His (rs11547464)"
)
R_VARS <- c("R151C", "R160W", "D294H", "D84E", "R142H")
r_VARS <- c("R163Q", "V60L", "V92M")
PEN_LEVELS <- c("r variant", "R variant", "R+r variant")

NOM_P  <- 0.05
MIN_N  <- 10
```

```{r helpers}
pct_diff <- function(bd, ref) {
  if (is.na(bd) || is.na(ref) || ref == 0) return(NA_real_)
  round(100 * bd / abs(ref), 1)
}

fmt_p <- function(p) {
  if (is.na(p)) return("")
  if (p < 0.001) return("<0.001")
  sprintf("%.3f", p)
}

# LME with fallback: varIdent can fail with unbalanced groups
safe_lme <- function(formula, data, random, weights = NULL, ...) {
  ctrl <- list(msMaxIter = 1000, msMaxEval = 1000, opt = "optim")
  tryCatch({
    if (!is.null(weights))
      mod <- lme(fixed = formula, random = random, data = data,
                 weights = weights, control = ctrl, ...)
    else
      mod <- lme(fixed = formula, random = random, data = data,
                 control = ctrl, ...)
    list(model = mod, success = TRUE, error = NULL)
  }, error = function(e) {
    tryCatch({
      mod <- lme(fixed = formula, random = random, data = data,
                 control = ctrl, ...)
      list(model = mod, success = TRUE, error = NULL,
           note = "Simplified (no heteroscedastic variance)")
    }, error = function(e2) list(model = NULL, success = FALSE, error = e2$message))
  })
}

# pull a named term from tTable
pull_term <- function(mod, term) {
  tt <- summary(mod)$tTable
  if (!term %in% rownames(tt)) return(NULL)
  list(b = tt[term, "Value"], se = tt[term, "Std.Error"], p = tt[term, "p-value"])
}

# extract group slopes from Cohort-interaction LME
extract_slopes <- function(mr, ref, data) {
  if (!mr$success) {
    return(data.frame(Group = NA, N = NA, Slope = NA, SE = NA, CI_lower = NA,
                      CI_upper = NA, Beta_diff = NA, Beta_SE = NA,
                      Beta_CI_lower = NA, Beta_CI_upper = NA,
                      Pct_diff = NA, P_value = NA, Error = mr$error))
  }
  mod <- mr$model
  fe <- fixef(mod); vm <- vcov(mod); fe_se <- sqrt(diag(vm))
  ns <- data %>% group_by(Cohort) %>%
    summarise(N = n_distinct(participant_id), .groups = "drop")
  ref_slope <- fe["visit_year"]; ref_se <- fe_se["visit_year"]

  lapply(levels(data$Cohort), function(coh) {
    n <- ns$N[ns$Cohort == coh]
    if (length(n) == 0) n <- NA

    if (coh == ref) {
      data.frame(Group = coh, N = n, Slope = ref_slope, SE = ref_se,
                 CI_lower = ref_slope - 1.96 * ref_se,
                 CI_upper = ref_slope + 1.96 * ref_se,
                 Beta_diff = NA_real_, Beta_SE = NA_real_,
                 Beta_CI_lower = NA_real_, Beta_CI_upper = NA_real_,
                 Pct_diff = NA_real_, P_value = NA_real_)
    } else {
      # nlme mangles spaces/+ in cohort names
      cc <- gsub(" ", ".", coh); cc <- gsub("\\+", ".", cc)
      nms <- c(paste0("visit_year:Cohort", coh), paste0("visit_year:Cohort", cc),
               paste0("Cohort", coh, ":visit_year"), paste0("Cohort", cc, ":visit_year"))
      found <- NULL
      for (nm in nms) if (nm %in% names(fe)) { found <- nm; break }

      if (!is.null(found)) {
        bd <- fe[found]; bse <- fe_se[found]
        slope <- ref_slope + bd
        var_s <- vm["visit_year","visit_year"] + vm[found,found] + 2*vm["visit_year",found]
        slope_se <- sqrt(max(0, var_s))
        pv <- tryCatch(summary(mod)$tTable[found, "p-value"], error = function(e) NA_real_)
        data.frame(Group = coh, N = n, Slope = slope, SE = slope_se,
                   CI_lower = slope - 1.96 * slope_se,
                   CI_upper = slope + 1.96 * slope_se,
                   Beta_diff = bd, Beta_SE = bse,
                   Beta_CI_lower = bd - 1.96 * bse,
                   Beta_CI_upper = bd + 1.96 * bse,
                   Pct_diff = pct_diff(bd, ref_slope), P_value = pv)
      } else {
        data.frame(Group = coh, N = n, Slope = NA, SE = NA, CI_lower = NA,
                   CI_upper = NA, Beta_diff = NA, Beta_SE = NA,
                   Beta_CI_lower = NA, Beta_CI_upper = NA,
                   Pct_diff = NA, P_value = NA)
      }
    }
  }) %>% do.call(rbind, .)
}

run_progression <- function(data, outcome, comp, ledd = FALSE, bl_col = NULL) {
  ad <- data %>%
    filter(Cohort %in% c(comp$ref, comp$test)) %>%
    mutate(Cohort = factor(Cohort, levels = c(comp$ref, comp$test)))

  if (is.null(bl_col)) {
    ad <- ad %>% group_by(participant_id) %>%
      mutate(baseline_score = first(!!sym(outcome), order_by = visit_year)) %>%
      ungroup()
    bl_col <- "baseline_score"
  }

  req <- c(outcome, "visit_year", bl_col, "Age_onset", "Sex", "Race",
           "Cohort", "participant_id")
  if (ledd) req <- c(req, "LEDD")
  ad <- ad %>% filter(if_all(all_of(req), ~ !is.na(.)))

  if (nrow(ad) < 20)
    return(list(success = FALSE,
                error = sprintf("Insufficient data: %d obs", nrow(ad)),
                comparison = comp))

  covs <- if (ledd) paste(bl_col, "+ Age_onset + Sex + Race + LEDD")
          else paste(bl_col, "+ Age_onset + Sex + Race")
  fml <- as.formula(sprintf("%s ~ visit_year * Cohort + %s", outcome, covs))

  mr <- safe_lme(fml, ad, ~ visit_year + 1 | participant_id,
                 weights = varIdent(form = ~ 1 | Cohort))
  slopes <- extract_slopes(mr, comp$ref, ad)
  slopes$Comparison <- paste(comp$test, "vs", comp$ref)
  slopes$Include_LEDD <- ledd

  list(success = mr$success, model = mr$model, slopes = slopes, comparison = comp,
       n_obs = nrow(ad), n_participants = n_distinct(ad$participant_id),
       small_sample_warning = if (nrow(ad) < 50) sprintf("Small sample (%d obs)", nrow(ad)))
}

# shared date/visit_year calculation for PPMI raw CSVs
load_ppmi_base <- function(path, valid_ids) {
  raw <- read_csv(path, show_col_types = FALSE)

  bl <- raw %>%
    filter(EVENT_ID == "BL", INFODT != "") %>%
    group_by(PATNO) %>% slice(1) %>% ungroup() %>%
    mutate(bl_date = as.Date(paste0("01/", INFODT), format = "%d/%m/%Y")) %>%
    select(PATNO, bl_date)

  raw %>%
    left_join(bl, by = "PATNO") %>%
    mutate(participant_id = paste0("PP-", PATNO),
           visit_date = as.Date(paste0("01/", INFODT), format = "%d/%m/%Y"),
           visit_year = as.numeric(difftime(visit_date, bl_date, units = "days")) / 365.25) %>%
    filter(!is.na(visit_year), visit_year >= 0, participant_id %in% valid_ids)
}

# row-level sum with minimum non-NA threshold
row_sum_threshold <- function(df, items, min_valid) {
  mat <- as.matrix(df[, items, drop = FALSE])
  storage.mode(mat) <- "double"
  valid <- rowSums(!is.na(mat))
  sums  <- rowSums(mat, na.rm = TRUE)
  ifelse(valid >= min_valid, sums, NA_real_)
}

# baseline cross-sectional biomarker comparison
bl_biomarker <- function(data, outcome, comp) {
  ad <- data %>% filter(Cohort %in% c(comp$ref, comp$test), !is.na(!!sym(outcome)))
  rv <- ad %>% filter(Cohort == comp$ref) %>% pull(!!sym(outcome))
  tv <- ad %>% filter(Cohort == comp$test) %>% pull(!!sym(outcome))
  nr <- length(rv); nt <- length(tv)
  lbl <- paste(comp$test, "vs", comp$ref)

  if (nr < 5 || nt < 5)
    return(data.frame(Comparison = lbl, Outcome = outcome, N_ref = nr, N_test = nt,
                      Mean_ref = NA, SD_ref = NA, Mean_test = NA, SD_test = NA,
                      Diff = NA, Pct_diff = NA, P_value_unadj = NA,
                      Beta_adj = NA, SE_adj = NA, CI_lower_adj = NA, CI_upper_adj = NA,
                      P_value_adj = NA, Analysis = "Combined"))

  mr <- mean(rv, na.rm = TRUE); mt <- mean(tv, na.rm = TRUE)
  unadj_p <- wilcox.test(tv, rv)$p.value

  # adjusted
  adj <- ad %>%
    mutate(Cohort = factor(Cohort, levels = c(comp$ref, comp$test))) %>%
    filter(!is.na(Age_onset), !is.na(Sex), !is.na(Race))
  adj_out <- tryCatch({
    m <- lm(as.formula(sprintf("%s ~ Cohort + Age_onset + Sex + Race", outcome)), data = adj)
    cn <- paste0("Cohort", comp$test)
    ms <- summary(m); ci <- confint(m, cn)
    if (cn %in% rownames(ms$coefficients))
      list(b = ms$coefficients[cn, "Estimate"], se = ms$coefficients[cn, "Std. Error"],
           lo = ci[1], hi = ci[2], p = ms$coefficients[cn, "Pr(>|t|)"])
    else list(b = NA, se = NA, lo = NA, hi = NA, p = NA)
  }, error = function(e) list(b = NA, se = NA, lo = NA, hi = NA, p = NA))

  data.frame(Comparison = lbl, Outcome = outcome, N_ref = nr, N_test = nt,
             Mean_ref = round(mr, 2), SD_ref = round(sd(rv, na.rm = TRUE), 2),
             Mean_test = round(mt, 2), SD_test = round(sd(tv, na.rm = TRUE), 2),
             Diff = round(mt - mr, 2), Pct_diff = round(100 * (mt - mr) / mr, 1),
             P_value_unadj = unadj_p,
             Beta_adj = round(adj_out$b, 3), SE_adj = round(adj_out$se, 3),
             CI_lower_adj = round(adj_out$lo, 3), CI_upper_adj = round(adj_out$hi, 3),
             P_value_adj = adj_out$p, Analysis = "Combined")
}
```

## Demographics

```{r demographics}
dem_raw <- read_excel(config$data, sheet = SH$demographics)

dem <- dem_raw %>%
  filter(Cohort %in% VALID_COHORTS) %>%
  mutate(
    Cohort    = factor(Cohort, levels = VALID_COHORTS),
    Sex       = factor(Sex, levels = c("Female", "Male")),
    Race      = factor(if_else(Race == "White", "White", "Other"), levels = c("White", "Other")),
    Ethnicity = factor(if_else(Ethnicity == "Hispanic or Latino", "Hispanic or Latino", "Other"),
                       levels = c("Hispanic or Latino", "Other"))
  )

dem <- dem %>%
  rowwise() %>%
  mutate(
    has_R = any(c_across(all_of(unname(VARIANT_COLS[R_VARS]))) > 0, na.rm = TRUE),
    has_r = any(c_across(all_of(unname(VARIANT_COLS[r_VARS]))) > 0, na.rm = TRUE),
    Penetrance = case_when(
      Cohort %in% c("SPD", "LRRK2 PD", "GBA PD") ~ "Non-carrier",
      has_R & has_r ~ "R+r variant",
      has_R ~ "R variant",
      has_r ~ "r variant",
      TRUE ~ "Unknown"
    ),
    # inclusive: R+r â†’ R (for stratification where R takes priority)
    Pen_inc = case_when(
      Cohort %in% c("SPD", "LRRK2 PD", "GBA PD") ~ "Non-carrier",
      has_R ~ "R variant",
      has_r ~ "r variant",
      TRUE ~ "Unknown"
    ),
    Genotype = case_when(
      Cohort %in% c("SPD", "LRRK2 PD", "GBA PD") ~ "Non-carrier",
      Homozygote == "Yes" ~ "Homozygote",
      TRUE ~ "Heterozygote"
    )
  ) %>% ungroup()

for (vn in names(VARIANT_COLS)) {
  col <- VARIANT_COLS[vn]
  if (col %in% names(dem)) dem[[paste0(vn, "_carrier")]] <- dem[[col]] > 0
}

dem <- dem %>%
  rename(Subject_ID   = `Subject ID`,
         Age_baseline = `Age at baseline`,
         Age_onset    = `Age at PD onset`,
         Diagnosis    = `Time since original diagnosis at baseline`)
dem$Age_baseline <- as.numeric(dem$Age_baseline)
dem$Age_onset    <- as.numeric(dem$Age_onset)
dem$Diagnosis    <- as.numeric(dem$Diagnosis)

VALID_IDS <- dem$Subject_ID
```

## Clinical Outcomes

```{r outcomes}
load_outcome <- function(sheet, score_col) {
  carrier_cols <- grep("_carrier$", names(dem), value = TRUE)
  extra <- c("Penetrance", "Pen_inc", "Genotype", carrier_cols)
  dem_cols <- c("Cohort", "Age_baseline", "Age_onset", "Diagnosis",
                "Sex", "Race", extra)
  raw <- read_excel(config$data, sheet = sheet)
  df <- raw %>%
    select(-any_of(dem_cols)) %>%
    left_join(dem %>% select(Subject_ID, all_of(dem_cols)),
              by = c("participant_id" = "Subject_ID")) %>%
    filter(!is.na(Cohort)) %>%
    group_by(participant_id) %>%
    mutate(baseline_score = first(!!sym(score_col), order_by = visit_year)) %>%
    ungroup() %>%
    rename(score = !!sym(score_col))
  df
}

d_u3     <- load_outcome(SH$updrs3, "mds_updrs_part_iii_summary_score")
d_u3_on  <- load_outcome(SH$updrs3_on, "mds_updrs_part_iii_summary_score")
d_u1     <- load_outcome(SH$updrs1, "mds_updrs_part_i_summary_score")
d_moca   <- load_outcome(SH$moca, "MoCA_score")

# UPDRS II/IV need EVENT_ID â†’ visit_year mapping
load_ppmi_part <- function(sheet, score_fn, label) {
  raw <- read_excel(config$data, sheet = sheet)
  raw$participant_id <- paste0("PP-", raw$PATNO)
  raw <- raw %>% filter(participant_id %in% VALID_IDS) %>%
    mutate(
      visit_month = case_when(
        EVENT_ID == "BL" ~ 0, EVENT_ID == "SC" ~ -1,
        grepl("^V", EVENT_ID) ~ as.numeric(gsub("V0?", "", EVENT_ID)) * 3,
        grepl("^R", EVENT_ID) ~ as.numeric(gsub("R0?", "", EVENT_ID)) * 6 + 48,
        TRUE ~ NA_real_),
      visit_year = visit_month / 12) %>%
    filter(!is.na(visit_year), visit_year >= 0)
  dem_cols <- c("Cohort", "Age_baseline", "Age_onset", "Diagnosis",
                "Sex", "Race", "Penetrance", "Pen_inc", "Genotype")
  carrier_cols <- grep("_carrier$", names(dem), value = TRUE)
  raw$score <- score_fn(raw)
  raw %>%
    select(-any_of(c(dem_cols, carrier_cols))) %>%
    left_join(dem %>% select(Subject_ID, all_of(dem_cols), all_of(carrier_cols)),
              by = c("participant_id" = "Subject_ID")) %>%
    filter(!is.na(Cohort), !is.na(score)) %>%
    group_by(participant_id) %>%
    mutate(baseline_score = first(score, order_by = visit_year)) %>%
    ungroup()
}

d_u2 <- load_ppmi_part(SH$updrs2, function(d) {
  if ("NP2PTOT" %in% names(d)) as.numeric(d$NP2PTOT)
  else rowSums(d[, grep("^NP2", names(d), value = TRUE)], na.rm = TRUE)
}, "UPDRS II")

d_u4 <- load_ppmi_part(SH$updrs4, function(d) as.numeric(d$NP4TOT), "UPDRS IV")

# DAT-SPECT
d_dat <- read_excel(config$data, sheet = SH$dat) %>%
  select(-any_of(c("Cohort","Age_baseline","Age_onset","Diagnosis",
                    "Sex","Race","Penetrance","Pen_inc","Genotype"))) %>%
  left_join(dem %>% select(Subject_ID, Cohort, Age_baseline, Age_onset, Diagnosis,
                           Sex, Race, Penetrance, Pen_inc, Genotype, ends_with("_carrier")),
            by = c("participant_id" = "Subject_ID")) %>%
  filter(!is.na(Cohort))

putamen_col <- grep("sbr_putamen", names(d_dat), value = TRUE)[1]
if (!is.na(putamen_col)) {
  d_dat <- d_dat %>% mutate(sbr_caudate = as.numeric(sbr_caudate),
                             sbr_putamen = as.numeric(.data[[putamen_col]]))
}
for (col in c("sbr_caudate_i", "sbr_caudate_c", "sbr_putamen_i", "sbr_putamen_c"))
  if (col %in% names(d_dat)) d_dat[[col]] <- as.numeric(d_dat[[col]])

d_dat <- d_dat %>%
  mutate(
    ratio_ci_put  = sbr_putamen_c / sbr_putamen_i,
    ratio_ci_cau  = sbr_caudate_c / sbr_caudate_i,
    ratio_pc_ipsi   = sbr_putamen_i / sbr_caudate_i,
    ratio_pc_contra = sbr_putamen_c / sbr_caudate_c,
    ratio_pc_mean   = sbr_putamen / sbr_caudate
  ) %>%
  group_by(participant_id) %>%
  mutate(bl_caudate = first(sbr_caudate, order_by = visit_year),
         bl_putamen = first(sbr_putamen, order_by = visit_year),
         bl_ci_put  = first(ratio_ci_put, order_by = visit_year),
         bl_ci_cau  = first(ratio_ci_cau, order_by = visit_year),
         bl_pc_ipsi = first(ratio_pc_ipsi, order_by = visit_year),
         bl_pc_contra = first(ratio_pc_contra, order_by = visit_year),
         bl_pc_mean = first(ratio_pc_mean, order_by = visit_year)) %>%
  ungroup()
```

```{r combined-updrs}
rnd_q <- function(x) round(x * 4) / 4

u1m <- d_u1 %>% mutate(vyq = rnd_q(visit_year)) %>%
  select(participant_id, vyq, score) %>% rename(u1 = score)
u2m <- d_u2 %>% mutate(vyq = rnd_q(visit_year)) %>%
  select(participant_id, vyq, score) %>% rename(u2 = score)
u3m <- d_u3 %>% mutate(vyq = rnd_q(visit_year)) %>%
  select(participant_id, visit_year, vyq, score, LEDD, Cohort, Age_onset,
         Sex, Race, Penetrance, Pen_inc, Genotype, ends_with("_carrier")) %>%
  rename(u3 = score)

d_comb <- u3m %>%
  left_join(u1m, by = c("participant_id", "vyq")) %>%
  left_join(u2m, by = c("participant_id", "vyq")) %>%
  filter(!is.na(u1), !is.na(u2)) %>%
  mutate(score = u1 + u2 + u3) %>%
  group_by(participant_id) %>%
  mutate(baseline_score = first(score, order_by = visit_year)) %>%
  ungroup()

u4m <- d_u4 %>% mutate(vyq = rnd_q(visit_year)) %>%
  select(participant_id, vyq, score) %>% rename(u4 = score)

d_total <- d_comb %>%
  left_join(u4m, by = c("participant_id", "vyq")) %>%
  filter(!is.na(u4)) %>%
  mutate(score = u1 + u2 + u3 + u4) %>%
  group_by(participant_id) %>%
  mutate(baseline_score = first(score, order_by = visit_year)) %>%
  ungroup()

TOTAL_OK <- nrow(d_total) > 100
```

## Primary: MDS-UPDRS III

```{r primary-updrs3}
u3_results <- list()
for (cn in names(COMPARISONS)) {
  comp <- COMPARISONS[[cn]]
  for (ledd in c(TRUE, FALSE)) {
    tag <- paste0(cn, if (ledd) "_ledd" else "_noledd")
    r <- run_progression(d_u3, "score", comp, ledd = ledd, bl_col = "baseline_score")
    if (r$success) {
      u3_results[[tag]] <- r$slopes %>%
        mutate(LEDD_adjusted = if (ledd) "Yes" else "No", Analysis_type = cn)
    }
  }
}
u3_df <- bind_rows(u3_results)
print(u3_df %>% select(Comparison, Group, N, Slope, Pct_diff, P_value, LEDD_adjusted) %>%
        mutate(Slope = round(Slope, 2), P_value = sapply(P_value, fmt_p)))
```

```{r primary-updrs3-on}
u3on_results <- list()
for (cn in names(COMPARISONS)) {
  comp <- COMPARISONS[[cn]]
  for (ledd in c(TRUE, FALSE)) {
    tag <- paste0(cn, if (ledd) "_ledd" else "_noledd")
    r <- run_progression(d_u3_on, "score", comp, ledd = ledd, bl_col = "baseline_score")
    if (r$success) {
      u3on_results[[tag]] <- r$slopes %>%
        mutate(LEDD_adjusted = if (ledd) "Yes" else "No", Analysis_type = cn)
    }
  }
}
u3on_df <- bind_rows(u3on_results)
print(u3on_df %>% select(Comparison, Group, N, Slope, Pct_diff, P_value, LEDD_adjusted) %>%
        mutate(Slope = round(Slope, 2), P_value = sapply(P_value, fmt_p)))
```

```{r bivariate-off-on}
# bivariate model stacking OFF and ON to formally test medication attenuation
# single model with state Ă— time interaction

# baseline OFF score as common anchor
bl_off <- d_u3 %>%
  filter(Cohort %in% c("SPD", "MC1R PD")) %>%
  group_by(participant_id) %>%
  summarise(bl_off = first(score, order_by = visit_year), .groups = "drop")

bv <- bind_rows(
  d_u3 %>% mutate(state = "OFF"),
  d_u3_on %>% mutate(state = "ON")
) %>%
  filter(Cohort %in% c("SPD", "MC1R PD")) %>%
  mutate(
    Cohort = factor(Cohort, levels = c("SPD", "MC1R PD")),
    state = factor(state, levels = c("OFF", "ON"))
  ) %>%
  left_join(bl_off, by = "participant_id") %>%
  filter(!is.na(score), !is.na(bl_off), !is.na(Age_onset),
         !is.na(Sex), !is.na(Race), !is.na(LEDD))

# full three-way model: state Ă— Cohort Ă— time
m_bv <- tryCatch({
  lme(score ~ visit_year * Cohort * state + bl_off + Age_onset + Sex + Race + LEDD,
      data = bv, random = ~ visit_year | participant_id,
      weights = varIdent(form = ~ 1 | state),
      control = lmeControl(opt = "optim", maxIter = 500, msMaxIter = 500),
      na.action = na.omit)
}, error = function(e) {
  # fallback without varIdent on state
  tryCatch({
    lme(score ~ visit_year * Cohort * state + bl_off + Age_onset + Sex + Race + LEDD,
        data = bv, random = ~ visit_year | participant_id,
        control = lmeControl(opt = "optim", maxIter = 500, msMaxIter = 500),
        na.action = na.omit)
  }, error = function(e2) NULL)
})

if (!is.null(m_bv)) {
  tt_bv <- summary(m_bv)$tTable

  cat("\nBivariate model, key coefficients:\n")
  key_rows <- grep("visit_year|state", rownames(tt_bv))
  print(round(tt_bv[key_rows, , drop = FALSE], 4))

  # extract slopes per group Ă— state
  fe <- fixef(m_bv)
  int_vy_coh <- fe["visit_year:CohortMC1R PD"]
  int_vy_st  <- fe["visit_year:stateON"]
  int_3way   <- if ("visit_year:CohortMC1R PD:stateON" %in% names(fe))
                  fe["visit_year:CohortMC1R PD:stateON"] else 0
  slope_spd_off  <- fe["visit_year"]
  slope_mc1r_off <- fe["visit_year"] + int_vy_coh
  slope_spd_on   <- fe["visit_year"] + int_vy_st
  slope_mc1r_on  <- fe["visit_year"] + int_vy_coh + int_vy_st + int_3way

  cat("\nEstimated slopes (pts/yr):\n")
  cat(sprintf("  Non-carriers OFF: %.3f\n", slope_spd_off))
  cat(sprintf("  Non-carriers ON:  %.3f\n", slope_spd_on))
  cat(sprintf("  MC1R carriers OFF: %.3f\n", slope_mc1r_off))
  cat(sprintf("  MC1R carriers ON:  %.3f\n", slope_mc1r_on))

  # state Ă— time p-value (medication attenuation)
  st_row <- which(rownames(tt_bv) == "visit_year:stateON")
  cat(sprintf("\n  State x time (medication attenuation): beta=%.3f, p=%.4f\n",
              tt_bv[st_row, "Value"], tt_bv[st_row, "p-value"]))

  # three-way interaction: does attenuation differ by MC1R?
  tw_row <- which(rownames(tt_bv) == "visit_year:CohortMC1R PD:stateON")
  if (length(tw_row) > 0) {
    cat(sprintf("  Three-way (MC1R x state x time): beta=%.3f, p=%.4f\n",
                tt_bv[tw_row, "Value"], tt_bv[tw_row, "p-value"]))
  }
} else {
  cat("Bivariate model failed to converge.\n")
}
```

```{r subscores}
sub_vars <- c("tremor_score", "rigidity_score", "bradykinesia_score",
              "axial_score", "bulbar_score", "pigd_score")

# merge subscore columns onto d_u3 / d_u3_on
merge_subscores <- function(main, sub_sheet) {
  subs <- read_excel(config$data, sheet = sub_sheet) %>%
    rename(score = mds_updrs_part_iii_summary_score) %>%
    arrange(participant_id, visit_year, score) %>%
    group_by(participant_id, visit_year, score) %>%
    mutate(dup_rank = row_number()) %>% ungroup()

  main <- main %>%
    arrange(participant_id, visit_year, score) %>%
    group_by(participant_id, visit_year, score) %>%
    mutate(dup_rank = row_number()) %>% ungroup()

  main <- main %>%
    left_join(subs %>% select(participant_id, visit_year, score, dup_rank, all_of(sub_vars)),
              by = c("participant_id", "visit_year", "score", "dup_rank"))
  main$dup_rank <- NULL
  main
}

d_u3    <- merge_subscores(d_u3, SH$updrs3_off_sub)
d_u3_on <- merge_subscores(d_u3_on, SH$updrs3_on_sub)

run_subscore <- function(data, sub_vars, state, ledd = FALSE,
                          comp = list(test = "MC1R PD", ref = "SPD")) {
  results <- list()
  for (sub in sub_vars) {
    lbl <- gsub("_score", "", sub)
    bl <- paste0("bl_", sub)

    sd <- data %>%
      filter(Cohort %in% c(comp$ref, comp$test), !is.na(.data[[sub]])) %>%
      mutate(Cohort = factor(Cohort, levels = c(comp$ref, comp$test))) %>%
      group_by(participant_id) %>%
      mutate(!!bl := first(.data[[sub]], order_by = visit_year)) %>%
      ungroup()

    covs <- paste(bl, "+ Age_onset + Sex + Race")
    if (ledd) covs <- paste(covs, "+ LEDD")
    req <- c(sub, "visit_year", bl, "Age_onset", "Sex", "Race", "Cohort", "participant_id")
    if (ledd) req <- c(req, "LEDD")
    sd <- sd %>% filter(if_all(all_of(req), ~ !is.na(.)))

    fml <- as.formula(paste0(sub, " ~ visit_year * Cohort + ", covs))

    fit <- tryCatch({
      lme(fml, random = ~ visit_year | participant_id, data = sd,
          control = lmeControl(opt = "optim", maxIter = 500, msMaxIter = 500))
    }, error = function(e) {
      # random slope can fail for low-variance subscores (tremor, bulbar)
      tryCatch(
        lme(fml, random = ~ 1 | participant_id, data = sd,
            control = lmeControl(opt = "optim")),
        error = function(e2) NULL)
    })

    if (!is.null(fit)) {
      fe <- fixef(fit); vm <- vcov(fit); fe_se <- sqrt(diag(vm))
      cs <- summary(fit)$tTable
      int_nm <- grep("visit_year:Cohort|Cohort.*:visit_year", names(fe), value = TRUE)

      if (length(int_nm) == 1) {
        rs <- fe["visit_year"]; rs_se <- fe_se["visit_year"]
        bd <- fe[int_nm]; bse <- fe_se[int_nm]
        ts <- rs + bd
        var_t <- vm["visit_year","visit_year"] + vm[int_nm,int_nm] + 2*vm["visit_year",int_nm]
        ts_se <- sqrt(max(0, var_t))

        results[[paste0(lbl, "_", ledd)]] <- data.frame(
          Subscore = lbl, Comparison = paste(comp$test, "vs", comp$ref),
          State = state, LEDD_adjusted = ledd,
          N_ref  = n_distinct(sd$participant_id[sd$Cohort == comp$ref]),
          N_test = n_distinct(sd$participant_id[sd$Cohort == comp$test]),
          N_obs  = nobs(fit),
          Ref_slope = rs, Ref_SE = rs_se,
          Ref_CI_lower = rs - 1.96*rs_se, Ref_CI_upper = rs + 1.96*rs_se,
          Test_slope = ts, Test_SE = ts_se,
          Test_CI_lower = ts - 1.96*ts_se, Test_CI_upper = ts + 1.96*ts_se,
          Diff = bd, Diff_SE = bse,
          Diff_CI_lower = bd - 1.96*bse, Diff_CI_upper = bd + 1.96*bse,
          Pct_diff = pct_diff(bd, rs), P_value = cs[int_nm, "p-value"])
      }
    }
  }
  if (length(results) > 0) do.call(rbind, results)
  else data.frame()
}

sub_off <- sub_on <- list()
for (cn in names(COMPARISONS)) {
  comp <- COMPARISONS[[cn]]
  for (ledd in c(FALSE, TRUE)) {
    tag <- paste0(cn, "_", ledd)
    r <- run_subscore(d_u3, sub_vars, "OFF", ledd, comp)
    if (nrow(r) > 0) { r$Analysis_type <- cn; sub_off[[tag]] <- r }
    r <- run_subscore(d_u3_on, sub_vars, "ON", ledd, comp)
    if (nrow(r) > 0) { r$Analysis_type <- cn; sub_on[[tag]] <- r }
  }
}
sub_off_df <- bind_rows(sub_off)
sub_on_df  <- bind_rows(sub_on)
sub_all_df <- bind_rows(sub_off_df, sub_on_df)

if (nrow(sub_off_df) > 0) {
  print(sub_off_df %>%
          mutate(across(where(is.numeric), ~ round(., 4))) %>%
          select(Comparison, Subscore, LEDD_adjusted, Ref_slope, Test_slope,
                 Diff, Pct_diff, P_value) %>%
          arrange(Comparison, LEDD_adjusted, P_value))
}
```

## Co-carrier Adjusted Models

```{r cocarrier-data}
make_cocarrier <- function(data) {
  data %>%
    mutate(
      MC1R_any  = as.integer(Cohort %in% c("MC1R PD", "MC1R + LRRK2 PD", "MC1R + GBA PD")),
      lrrk2_carrier = as.integer(Cohort %in% c("LRRK2 PD", "MC1R + LRRK2 PD")),
      gba_carrier   = as.integer(Cohort %in% c("GBA PD", "MC1R + GBA PD")),
      CC_Group  = factor(if_else(MC1R_any == 1, "MC1R", "Non-MC1R"),
                         levels = c("Non-MC1R", "MC1R")),
      # R+r compound carriers count toward both penetrance classes
      has_r = Pen_inc %in% "r variant" | Penetrance %in% c("r variant", "R+r variant"),
      has_R = Pen_inc %in% "R variant" | Penetrance %in% c("R variant", "R+r variant"),
      r_carrier = as.integer(MC1R_any == 1 & has_r),
      R_carrier = as.integer(MC1R_any == 1 & has_R)
    )
}

d_cc     <- make_cocarrier(d_u3)
d_cc_on  <- make_cocarrier(d_u3_on)

pts_cc <- d_cc %>% distinct(participant_id, .keep_all = TRUE)
cat(sprintf("Co-carrier: %d total (%d MC1R-any, %d non-MC1R, %d LRRK2, %d GBA)\n",
            nrow(pts_cc), sum(pts_cc$MC1R_any), sum(!pts_cc$MC1R_any),
            sum(pts_cc$lrrk2_carrier), sum(pts_cc$gba_carrier)))
cat(sprintf("  r-variant: %d | R-variant: %d\n", sum(pts_cc$r_carrier), sum(pts_cc$R_carrier)))
```

```{r cocarrier-models}
run_cc <- function(data, outcome, exposure, include_ledd = FALSE, label = "") {
  bl <- paste0("bl_", outcome)
  d <- data %>%
    filter(!is.na(.data[[outcome]])) %>%
    group_by(participant_id) %>%
    mutate(!!bl := first(.data[[outcome]], order_by = visit_year)) %>%
    ungroup()

  covs <- paste(bl, "+ Age_onset + Sex + Race + lrrk2_carrier + gba_carrier")
  if (include_ledd) covs <- paste(covs, "+ LEDD")
  req <- c(outcome, "visit_year", bl, "Age_onset", "Sex", "Race",
           "lrrk2_carrier", "gba_carrier", exposure, "participant_id")
  if (include_ledd) req <- c(req, "LEDD")
  d <- d %>% filter(if_all(all_of(req), ~ !is.na(.)))

  if (nrow(d) < 20 || n_distinct(d$participant_id) < 10) return(NULL)

  fml <- as.formula(sprintf("%s ~ visit_year * %s + %s", outcome, exposure, covs))
  fit <- tryCatch({
    lme(fml, random = ~ visit_year | participant_id, data = d,
        control = lmeControl(opt = "optim", maxIter = 500, msMaxIter = 500))
  }, error = function(e) {
    tryCatch(lme(fml, random = ~ 1 | participant_id, data = d,
                 control = lmeControl(opt = "optim")),
             error = function(e2) NULL)
  })
  if (is.null(fit)) return(NULL)

  fe <- fixef(fit); vm <- vcov(fit); fe_se <- sqrt(diag(vm))
  cs <- summary(fit)$tTable
  int_nm <- paste0("visit_year:", exposure)
  if (!int_nm %in% names(fe)) return(NULL)

  rs <- fe["visit_year"]; rs_se <- fe_se["visit_year"]
  bd <- fe[int_nm]; bse <- fe_se[int_nm]
  ts <- rs + bd
  var_t <- vm["visit_year","visit_year"] + vm[int_nm,int_nm] + 2*vm["visit_year",int_nm]
  ts_se <- sqrt(max(0, var_t))

  data.frame(
    Subscore = label, LEDD_adjusted = include_ledd, Exposure = exposure,
    N_exposed   = n_distinct(d$participant_id[d[[exposure]] == 1]),
    N_unexposed = n_distinct(d$participant_id[d[[exposure]] == 0]),
    N_obs = nobs(fit),
    Ref_slope = rs, Ref_SE = rs_se,
    Ref_CI_lower = rs - 1.96*rs_se, Ref_CI_upper = rs + 1.96*rs_se,
    Test_slope = ts, Test_SE = ts_se,
    Test_CI_lower = ts - 1.96*ts_se, Test_CI_upper = ts + 1.96*ts_se,
    Diff = bd, Diff_SE = bse,
    Diff_CI_lower = bd - 1.96*bse, Diff_CI_upper = bd + 1.96*bse,
    Pct_diff = pct_diff(bd, rs), P_value = cs[int_nm, "p-value"])
}

run_cc_suite <- function(data, exposure, sub_vars, sub_labels, state) {
  results <- list()
  for (i in seq_along(sub_vars)) {
    for (ledd in c(FALSE, TRUE)) {
      r <- run_cc(data, sub_vars[i], exposure, ledd, sub_labels[i])
      if (!is.null(r)) { r$State <- state; results[[paste0(sub_vars[i], "_", ledd)]] <- r }
    }
  }
  if (length(results) > 0) bind_rows(results) else data.frame()
}

cc_vars   <- c("score", "tremor_score", "rigidity_score",
               "bradykinesia_score", "axial_score", "bulbar_score")
cc_labels <- c("Total UPDRS III", "Tremor", "Rigidity",
               "Bradykinesia", "Axial", "Bulbar")

cc_off_full <- run_cc_suite(d_cc, "MC1R_any", cc_vars, cc_labels, "OFF")
cc_on_full  <- run_cc_suite(d_cc_on, "MC1R_any", cc_vars, cc_labels, "ON")

cc_off_r <- run_cc_suite(d_cc, "r_carrier", cc_vars, cc_labels, "OFF")
cc_off_R <- run_cc_suite(d_cc, "R_carrier", cc_vars, cc_labels, "OFF")
cc_on_r  <- run_cc_suite(d_cc_on, "r_carrier", cc_vars, cc_labels, "ON")
cc_on_R  <- run_cc_suite(d_cc_on, "R_carrier", cc_vars, cc_labels, "ON")

if (nrow(cc_off_r) > 0) cc_off_r$Penetrance <- "r-variant"
if (nrow(cc_off_R) > 0) cc_off_R$Penetrance <- "R-variant"
if (nrow(cc_on_r) > 0)  cc_on_r$Penetrance  <- "r-variant"
if (nrow(cc_on_R) > 0)  cc_on_R$Penetrance  <- "R-variant"

cc_pen_off <- bind_rows(cc_off_r, cc_off_R)
cc_pen_on  <- bind_rows(cc_on_r, cc_on_R)
cc_all <- bind_rows(
  cc_off_full %>% mutate(Penetrance = "All MC1R"),
  cc_on_full  %>% mutate(Penetrance = "All MC1R"),
  cc_pen_off, cc_pen_on)
```

## Secondary Outcomes

```{r secondary}
run_sec <- function(data, label, ledd = FALSE) {
  res <- lapply(names(COMPARISONS), function(cn) {
    r <- run_progression(data, "score", COMPARISONS[[cn]], ledd, "baseline_score")
    if (r$success) r$slopes %>% mutate(Outcome = label, Analysis_type = cn) else NULL
  }) %>% bind_rows()
  if (nrow(res) > 0)
    print(res %>% select(Outcome, Comparison, Group, N, Slope, Pct_diff, P_value) %>%
            mutate(Slope = round(Slope, 3), P_value = sapply(P_value, fmt_p)))
  res
}

u1_res   <- run_sec(d_u1, "MDS-UPDRS I")
moca_res <- run_sec(d_moca, "MoCA")

d_dat_c <- d_dat %>% rename(score = sbr_caudate, baseline_score = bl_caudate) %>% filter(!is.na(score))
d_dat_p <- d_dat %>% rename(score = sbr_putamen, baseline_score = bl_putamen) %>% filter(!is.na(score))
dat_c_res <- run_sec(d_dat_c, "DAT-SPECT Caudate")
dat_p_res <- run_sec(d_dat_p, "DAT-SPECT Putamen")

u2_res    <- run_sec(d_u2, "MDS-UPDRS II")
u4_res    <- run_sec(d_u4, "MDS-UPDRS IV")
comb_res  <- if (nrow(d_comb) > 100) run_sec(d_comb, "Combined MDS-UPDRS I-III", ledd = TRUE)
total_res <- if (TOTAL_OK) run_sec(d_total, "Total MDS-UPDRS I-IV", ledd = TRUE)
```

## CSF Biomarkers

```{r csf-biomarkers}
d_saa <- read_excel(config$data, sheet = SH$saa_detailed) %>%
  filter(Cohort %in% VALID_COHORTS) %>%
  mutate(
    Cohort    = factor(Cohort, levels = VALID_COHORTS),
    Sex       = factor(Sex, levels = c("Female", "Male")),
    Race      = factor(if_else(Race == "White", "White", "Other"), levels = c("White", "Other")),
    Ethnicity = factor(if_else(Ethnicity == "Hispanic or Latino", "Hispanic or Latino", "Other"),
                       levels = c("Hispanic or Latino", "Other"))
  ) %>%
  rowwise() %>%
  mutate(
    Fmax_avg = mean(c(Fmax_150h_Rep1, Fmax_150h_Rep2, Fmax_150h_Rep3), na.rm = TRUE),
    AUC_avg  = mean(c(AUC_150h_Rep1, AUC_150h_Rep2, AUC_150h_Rep3), na.rm = TRUE),
    TTT_avg  = mean(c(TTT_150h_Rep1, TTT_150h_Rep2, TTT_150h_Rep3), na.rm = TRUE),
    T50_avg  = mean(c(T50_150h_Rep1, T50_150h_Rep2, T50_150h_Rep3), na.rm = TRUE),
    has_R = any(c_across(all_of(unname(VARIANT_COLS[R_VARS]))) > 0, na.rm = TRUE),
    has_r = any(c_across(all_of(unname(VARIANT_COLS[r_VARS]))) > 0, na.rm = TRUE),
    Penetrance = case_when(
      Cohort %in% c("SPD", "LRRK2 PD", "GBA PD") ~ "Non-carrier",
      has_R ~ "R variant", has_r ~ "r variant", TRUE ~ "Unknown"),
    Genotype = case_when(
      Cohort %in% c("SPD", "LRRK2 PD", "GBA PD") ~ "Non-carrier",
      Homozygote == 1 ~ "Homozygote", TRUE ~ "Heterozygote")
  ) %>% ungroup() %>%
  rename(Age_baseline = `Age at baseline`,
         Age_onset    = `Age at PD onset`,
         Diagnosis    = `Time since original diagnosis at baseline`)

saa_bl <- d_saa %>% filter(CLINICAL_EVENT == "BL")
saa_measures <- c("Fmax_avg", "AUC_avg", "TTT_avg", "T50_avg")

saa_bl_df <- lapply(saa_measures, function(m) {
  lapply(names(COMPARISONS), function(cn) {
    r <- bl_biomarker(saa_bl, m, COMPARISONS[[cn]]); r$Analysis <- cn; r
  }) %>% bind_rows()
}) %>% bind_rows()

# SAA by penetrance
saa_pen <- lapply(c("r variant", "R variant"), function(pen) {
  mc1r_sub <- saa_bl %>% filter(Cohort == "MC1R PD", Penetrance == pen)
  spd <- saa_bl %>% filter(Cohort == "SPD")
  if (nrow(mc1r_sub) < MIN_N) return(NULL)
  grp <- paste0(pen, " MC1R PD")
  sd <- bind_rows(
    spd %>% mutate(Cohort = factor("SPD", levels = c("SPD", grp))),
    mc1r_sub %>% mutate(Cohort = factor(grp, levels = c("SPD", grp))))
  lapply(saa_measures, function(m) {
    bl_biomarker(sd, m, list(ref = "SPD", test = grp)) %>% mutate(Penetrance = pen)
  }) %>% bind_rows()
}) %>% bind_rows()

# CSF Tau/pTau/Abeta
d_csf <- read_excel(config$data, sheet = SH$csf) %>%
  select(-any_of(c("Cohort","Age_onset","Sex","Race","Penetrance","Genotype"))) %>%
  left_join(dem %>% select(Subject_ID, Cohort, Age_onset, Sex, Race, Penetrance, Genotype,
                           ends_with("_carrier")),
            by = c("participant_id" = "Subject_ID")) %>%
  filter(!is.na(Cohort)) %>%
  mutate(visit_year = visit_month / 12)

d_csf_w <- d_csf %>%
  select(participant_id, Cohort, Age_onset, Sex, Race, Penetrance, Genotype,
         visit_year, test_name, test_value) %>%
  pivot_wider(names_from = test_name, values_from = test_value, values_fn = mean)

csf_bl <- d_csf_w %>% filter(visit_year == 0)
csf_bl_df <- lapply(c("Tau", "p-Tau", "Abeta"), function(m) {
  lapply(names(COMPARISONS), function(cn) {
    r <- bl_biomarker(csf_bl, m, COMPARISONS[[cn]]); r$Outcome <- m; r$Analysis <- cn; r
  }) %>% bind_rows()
}) %>% bind_rows()

# alpha-synuclein
d_asyn <- read_excel(config$data, sheet = SH$asyn) %>%
  select(-any_of(c("Cohort","Age_onset","Sex","Race","Penetrance","Genotype"))) %>%
  left_join(dem %>% select(Subject_ID, Cohort, Age_onset, Sex, Race, Penetrance, Genotype,
                           ends_with("_carrier")),
            by = c("participant_id" = "Subject_ID")) %>%
  filter(!is.na(Cohort)) %>%
  rename(visit_year = visit_year2) %>%
  mutate(score = as.numeric(Concentration)) %>%
  filter(!is.na(score)) %>%
  group_by(participant_id) %>%
  mutate(baseline_score = first(score, order_by = visit_year)) %>%
  ungroup()

asyn_bl <- lapply(names(COMPARISONS), function(cn) {
  comp <- COMPARISONS[[cn]]
  nt <- n_distinct(d_asyn$participant_id[d_asyn$Cohort == comp$test])
  nr <- n_distinct(d_asyn$participant_id[d_asyn$Cohort == comp$ref])
  if (nt >= 3 && nr >= 3) {
    r <- bl_biomarker(d_asyn, "score", comp)
    r$Outcome <- "Alpha-Synuclein"; r$Analysis <- cn; r
  } else NULL
}) %>% bind_rows()

# NfL / GFAP
d_nfl_gfap <- read_excel(config$data, sheet = SH$nfl_gfap) %>%
  filter(Cohort %in% VALID_COHORTS) %>%
  mutate(Cohort     = factor(Cohort, levels = VALID_COHORTS),
         Sex        = factor(SEX, levels = c("Female", "Male")),
         test_name  = if_else(TESTNAME %in% c("NFL", "NfL"), "NfL", "GFAP"),
         test_value = as.numeric(TESTVALUE))

# resolve ID column mismatch (PATNO vs PP-PATNO)
nfl_id_col <- grep("Subject|participant|PATNO", names(d_nfl_gfap), value = TRUE, ignore.case = TRUE)[1]
nfl_ids <- as.character(d_nfl_gfap[[nfl_id_col]])
if (sum(nfl_ids %in% dem$Subject_ID) == 0 && !any(grepl("^PP-", nfl_ids))) {
  d_nfl_gfap$pid_fmt <- paste0("PP-", d_nfl_gfap[[nfl_id_col]])
  if (sum(d_nfl_gfap$pid_fmt %in% dem$Subject_ID) > 0) nfl_id_col <- "pid_fmt"
}

# drop assembled demographics before joining canonical versions from dem
d_nfl_gfap <- d_nfl_gfap %>%
  select(-any_of(c("Age_onset", "Race", "Penetrance", "Pen_inc", "Genotype"))) %>%
  left_join(dem %>% select(Subject_ID, Age_onset, Race, Penetrance, Pen_inc, Genotype,
                           ends_with("_carrier")),
            by = setNames("Subject_ID", nfl_id_col))

# baseline filter
if ("EVENT_ID" %in% names(d_nfl_gfap)) {
  bl_filt <- d_nfl_gfap$EVENT_ID == "BL"
} else if ("CLINICAL_EVENT" %in% names(d_nfl_gfap)) {
  bl_filt <- grepl("Baseline|BL", d_nfl_gfap$CLINICAL_EVENT, ignore.case = TRUE)
} else if ("visit_year" %in% names(d_nfl_gfap)) {
  bl_filt <- d_nfl_gfap$visit_year == 0
} else {
  d_nfl_gfap <- d_nfl_gfap %>%
    group_by(.data[[nfl_id_col]], test_name) %>% mutate(is_first = row_number() == 1) %>% ungroup()
  bl_filt <- d_nfl_gfap$is_first
}

nfl_df <- lapply(names(COMPARISONS), function(cn) {
  r <- bl_biomarker(d_nfl_gfap %>% filter(bl_filt, !is.na(test_value), test_name == "NfL"),
                    "test_value", COMPARISONS[[cn]]); r$Outcome <- "NfL"; r
}) %>% bind_rows()

gfap_df <- lapply(names(COMPARISONS), function(cn) {
  r <- bl_biomarker(d_nfl_gfap %>% filter(bl_filt, !is.na(test_value), test_name == "GFAP"),
                    "test_value", COMPARISONS[[cn]]); r$Outcome <- "GFAP"; r
}) %>% bind_rows()
```

## Interactions

```{r interactions}
d_int <- d_u3 %>%
  filter(Cohort %in% c("SPD", "MC1R PD")) %>%
  mutate(MC1R = as.integer(Cohort == "MC1R PD"))

# Age x MC1R x Time
d_age <- d_int %>% mutate(Age_c = Age_onset - mean(Age_onset, na.rm = TRUE))
age_int <- tryCatch({
  m <- lme(score ~ visit_year * MC1R * Age_c + baseline_score + Sex + Race + LEDD,
           data = d_age, random = ~ visit_year + 1 | participant_id,
           control = lmeControl(maxIter = 1000, msMaxIter = 1000, opt = "optim"), na.action = na.omit)
  tt <- summary(m)$tTable
  if ("visit_year:MC1R:Age_c" %in% rownames(tt)) {
    data.frame(Analysis = "Age x MC1R x Time", Model = "Three-way",
               N = n_distinct(d_age$participant_id),
               Mean_age = round(mean(d_age$Age_onset, na.rm = TRUE), 1),
               MC1R_beta = round(tt["visit_year:MC1R","Value"], 4),
               MC1R_SE = round(tt["visit_year:MC1R","Std.Error"], 4),
               MC1R_p = tt["visit_year:MC1R","p-value"],
               Three_way_beta = round(tt["visit_year:MC1R:Age_c","Value"], 4),
               Three_way_SE = round(tt["visit_year:MC1R:Age_c","Std.Error"], 4),
               Three_way_p = tt["visit_year:MC1R:Age_c","p-value"])
  } else data.frame()
}, error = function(e) data.frame())
AGE_INT_OK <- nrow(age_int) > 0

# Sex x MC1R x Time
d_sex <- d_int %>% mutate(Sex_bin = as.integer(Sex == "Male"))
sex_int <- tryCatch({
  m <- lme(score ~ visit_year * MC1R * Sex_bin + baseline_score + Age_onset + Race + LEDD,
           data = d_sex, random = ~ visit_year + 1 | participant_id,
           control = lmeControl(maxIter = 1000, msMaxIter = 1000, opt = "optim"), na.action = na.omit)
  tt <- summary(m)$tTable
  if ("visit_year:MC1R:Sex_bin" %in% rownames(tt)) {
    p3 <- tt["visit_year:MC1R:Sex_bin","p-value"]
    data.frame(Analysis = "Sex x MC1R x Time", Model = "Three-way",
               N = n_distinct(d_sex$participant_id),
               N_male = n_distinct(d_sex$participant_id[d_sex$Sex == "Male"]),
               N_female = n_distinct(d_sex$participant_id[d_sex$Sex == "Female"]),
               MC1R_beta = round(tt["visit_year:MC1R","Value"], 4),
               MC1R_SE = round(tt["visit_year:MC1R","Std.Error"], 4),
               MC1R_p = tt["visit_year:MC1R","p-value"],
               Three_way_beta = round(tt["visit_year:MC1R:Sex_bin","Value"], 4),
               Three_way_SE = round(tt["visit_year:MC1R:Sex_bin","Std.Error"], 4),
               Three_way_p = p3)
  } else data.frame()
}, error = function(e) data.frame())
SEX_INT_OK <- nrow(sex_int) > 0

# Duration x MC1R x Time
d_dur <- d_int %>% filter(!is.na(Diagnosis)) %>%
  mutate(Dx_c = Diagnosis - mean(Diagnosis, na.rm = TRUE))
dur_int <- tryCatch({
  m <- lme(score ~ visit_year * MC1R * Dx_c + baseline_score + Age_onset + Sex + Race + LEDD,
           data = d_dur, random = ~ visit_year + 1 | participant_id,
           control = lmeControl(maxIter = 1000, msMaxIter = 1000, opt = "optim"), na.action = na.omit)
  tt <- summary(m)$tTable
  if ("visit_year:MC1R:Dx_c" %in% rownames(tt)) {
    data.frame(Analysis = "Duration x MC1R x Time", Model = "Three-way",
               N = n_distinct(d_dur$participant_id),
               Mean_duration = round(mean(d_dur$Diagnosis, na.rm = TRUE), 1),
               MC1R_beta = round(tt["visit_year:MC1R","Value"], 4),
               MC1R_SE = round(tt["visit_year:MC1R","Std.Error"], 4),
               MC1R_p = tt["visit_year:MC1R","p-value"],
               Three_way_beta = round(tt["visit_year:MC1R:Dx_c","Value"], 4),
               Three_way_SE = round(tt["visit_year:MC1R:Dx_c","Std.Error"], 4),
               Three_way_p = tt["visit_year:MC1R:Dx_c","p-value"])
  } else data.frame()
}, error = function(e) data.frame())
DUR_INT_OK <- nrow(dur_int) > 0
```

## Genetic Independence

Tests whether MC1R progression effect is independent of GBA/LRRK2 status.
Three-way interactions fit separately because no LRRK2+GBA co-carrier cohort
exists in PPMI (empty cell problem).

```{r gen-independence}
d_gi <- d_u3 %>%
  mutate(
    MC1R  = as.integer(Cohort %in% c("MC1R PD", "MC1R + GBA PD", "MC1R + LRRK2 PD")),
    GBA   = as.integer(Cohort %in% c("GBA PD", "MC1R + GBA PD")),
    LRRK2 = as.integer(Cohort %in% c("LRRK2 PD", "MC1R + LRRK2 PD")),
    gen_group = factor(case_when(
      MC1R == 1 & GBA == 1   ~ "MC1R+GBA",
      MC1R == 1 & LRRK2 == 1 ~ "MC1R+LRRK2",
      MC1R == 1               ~ "MC1R-only",
      GBA == 1                ~ "GBA-only",
      LRRK2 == 1              ~ "LRRK2-only",
      TRUE                    ~ "SPD"
    ), levels = c("SPD","MC1R-only","GBA-only","LRRK2-only","MC1R+GBA","MC1R+LRRK2"))
  ) %>%
  group_by(participant_id) %>%
  mutate(baseline_score = first(score, order_by = visit_year)) %>%
  ungroup() %>%
  filter(!is.na(score), !is.na(baseline_score),
         !is.na(Age_onset), !is.na(Sex), !is.na(Race), !is.na(LEDD))

print(d_gi %>% distinct(participant_id, gen_group) %>% count(gen_group))

# M1: MC1R x time only
gi_m1 <- safe_lme(score ~ visit_year * MC1R + baseline_score + Age_onset + Sex + Race + LEDD,
                   d_gi, ~ visit_year + 1 | participant_id,
                   weights = varIdent(form = ~ 1 | gen_group))

# M2: MC1R + GBA + LRRK2 mutually adjusted
gi_m2 <- safe_lme(score ~ visit_year * MC1R + visit_year * GBA + visit_year * LRRK2 +
                     baseline_score + Age_onset + Sex + Race + LEDD,
                   d_gi, ~ visit_year + 1 | participant_id,
                   weights = varIdent(form = ~ 1 | gen_group))

# M3a: MC1R x GBA x time
gi_m3a <- safe_lme(score ~ visit_year * MC1R * GBA + baseline_score + Age_onset + Sex + Race + LEDD,
                    d_gi, ~ visit_year + 1 | participant_id,
                    weights = varIdent(form = ~ 1 | gen_group))

# M3b: MC1R x LRRK2 x time
gi_m3b <- safe_lme(score ~ visit_year * MC1R * LRRK2 + baseline_score + Age_onset + Sex + Race + LEDD,
                    d_gi, ~ visit_year + 1 | participant_id,
                    weights = varIdent(form = ~ 1 | gen_group))

# coefficient stability table
gi_coef <- data.frame(Model = character(), Description = character(),
                       MC1R_beta = numeric(), MC1R_SE = numeric(),
                       MC1R_CI_lower = numeric(), MC1R_CI_upper = numeric(),
                       MC1R_p = numeric())

for (mi in list(
  list("M1",  "MC1R only",             gi_m1),
  list("M2",  "MC1R + GBA + LRRK2",    gi_m2),
  list("M3a", "MC1R x GBA",            gi_m3a),
  list("M3b", "MC1R x LRRK2",          gi_m3b)
)) {
  if (mi[[3]]$success) {
    t <- pull_term(mi[[3]]$model, "visit_year:MC1R")
    if (!is.null(t))
      gi_coef <- rbind(gi_coef, data.frame(
        Model = mi[[1]], Description = mi[[2]],
        MC1R_beta = round(t$b, 4), MC1R_SE = round(t$se, 4),
        MC1R_CI_lower = round(t$b - 1.96*t$se, 4),
        MC1R_CI_upper = round(t$b + 1.96*t$se, 4),
        MC1R_p = t$p))
  }
}
print(gi_coef)

# percent attenuation M1 -> M2
if (nrow(gi_coef) >= 2) {
  b1 <- gi_coef$MC1R_beta[gi_coef$Model == "M1"]
  b2 <- gi_coef$MC1R_beta[gi_coef$Model == "M2"]
  if (length(b1) == 1 && length(b2) == 1 && b1 != 0)
    cat(sprintf("M1 -> M2 attenuation: %.1f%%\n", 100 * (b2 - b1) / abs(b1)))
}

# within-stratum consistency
strata <- list(
  list("GBA-/LRRK2- (sporadic)", quote(GBA == 0 & LRRK2 == 0)),
  list("GBA carriers",           quote(GBA == 1)),
  list("LRRK2 carriers",         quote(LRRK2 == 1))
)

gi_strata <- lapply(strata, function(s) {
  ds <- d_gi %>% filter(!!s[[2]])
  nm <- n_distinct(ds$participant_id[ds$MC1R == 1])
  nr <- n_distinct(ds$participant_id[ds$MC1R == 0])
  if (nm < 5 || nr < 5)
    return(data.frame(Stratum = s[[1]], N_MC1R = nm, N_ref = nr,
                      Ref_slope = NA, Beta = NA, SE = NA,
                      CI_lower = NA, CI_upper = NA, Pct_diff = NA, P_value = NA))

  ds$MC1R_f <- factor(ds$MC1R, levels = c(0, 1), labels = c("Non-MC1R", "MC1R"))
  fit <- safe_lme(score ~ visit_year * MC1R + baseline_score + Age_onset + Sex + Race + LEDD,
                  ds, ~ visit_year + 1 | participant_id,
                  weights = varIdent(form = ~ 1 | MC1R_f))

  if (fit$success) {
    t <- pull_term(fit$model, "visit_year:MC1R")
    rs <- pull_term(fit$model, "visit_year")
    if (!is.null(t) && !is.null(rs))
      return(data.frame(Stratum = s[[1]], N_MC1R = nm, N_ref = nr,
                        Ref_slope = round(rs$b, 4),
                        Beta = round(t$b, 4), SE = round(t$se, 4),
                        CI_lower = round(t$b - 1.96*t$se, 4),
                        CI_upper = round(t$b + 1.96*t$se, 4),
                        Pct_diff = pct_diff(t$b, rs$b), P_value = t$p))
  }
  data.frame(Stratum = s[[1]], N_MC1R = nm, N_ref = nr,
             Ref_slope = NA, Beta = NA, SE = NA,
             CI_lower = NA, CI_upper = NA, Pct_diff = NA, P_value = NA)
}) %>% bind_rows()
print(gi_strata)

GEN_IND_OK <- nrow(gi_coef) > 0
```

## Sensitivity

```{r sensitivity}
COV_SETS <- list(
  full     = list(name = "Full (Age_onset + Sex + Race + LEDD)",
                  fml = "score ~ visit_year * Cohort + baseline_score + Age_onset + Sex + Race + LEDD",
                  req = c("score","baseline_score","Age_onset","Sex","Race","LEDD")),
  no_ledd  = list(name = "No LEDD",
                  fml = "score ~ visit_year * Cohort + baseline_score + Age_onset + Sex + Race",
                  req = c("score","baseline_score","Age_onset","Sex","Race")),
  minimal  = list(name = "Minimal (Age_onset + Sex)",
                  fml = "score ~ visit_year * Cohort + baseline_score + Age_onset + Sex",
                  req = c("score","baseline_score","Age_onset","Sex")),
  old_spec = list(name = "Age_baseline + Duration",
                  fml = "score ~ visit_year * Cohort + baseline_score + Age_baseline + Diagnosis + Sex + Race + LEDD",
                  req = c("score","baseline_score","Age_baseline","Diagnosis","Sex","Race","LEDD"))
)

sens_data <- d_u3 %>%
  filter(Cohort %in% c("SPD", "MC1R PD")) %>%
  mutate(Cohort = factor(Cohort, levels = c("SPD", "MC1R PD")))

# sample availability per covariate set
ss_table <- lapply(names(COV_SETS), function(sn) {
  si <- COV_SETS[[sn]]; ac <- si$req[si$req %in% names(sens_data)]
  cd <- sens_data %>% filter(complete.cases(across(all_of(ac))))
  data.frame(Set = si$name, N_obs = nrow(cd),
             N_participants = n_distinct(cd$participant_id),
             Pct_retained = round(100 * nrow(cd) / nrow(sens_data), 1))
}) %>% bind_rows()
print(ss_table)

sens_results <- list()
for (sn in names(COV_SETS)) {
  si <- COV_SETS[[sn]]; ac <- si$req[si$req %in% names(sens_data)]
  cd <- sens_data %>% filter(complete.cases(across(all_of(ac))))
  if (n_distinct(cd$participant_id) < 50) next
  tryCatch({
    mod <- lme(as.formula(si$fml), cd, ~ visit_year + 1 | participant_id,
               control = lmeControl(maxIter = 1000, msMaxIter = 1000, opt = "optim"),
               na.action = na.omit)
    tt <- summary(mod)$tTable
    ir <- grep("visit_year:Cohort", rownames(tt))
    if (length(ir) > 0)
      sens_results[[sn]] <- data.frame(
        Covariate_Set = si$name, N = n_distinct(cd$participant_id),
        Beta = round(tt[ir,"Value"], 3), SE = round(tt[ir,"Std.Error"], 3),
        P_value = tt[ir,"p-value"])
  }, error = function(e) NULL)
}
if (length(sens_results) > 0) print(bind_rows(sens_results))
```

```{r sensitivity-time-interactions}
# Sensitivity: covariate Ă— time interactions
# Test whether covariate Ă— time interactions change the primary result
# All covariates interact with time except LEDD (time-varying, interaction would be collinear)
ti_data <- d_u3 %>%
  filter(Cohort %in% c("SPD", "MC1R PD")) %>%
  mutate(Cohort = factor(Cohort, levels = c("SPD", "MC1R PD"))) %>%
  filter(!is.na(score), !is.na(baseline_score), !is.na(Age_onset),
         !is.na(Sex), !is.na(Race), !is.na(LEDD))

# current model (baseline adjustment only)
m_base <- safe_lme(
  score ~ visit_year * Cohort + baseline_score + Age_onset + Sex + Race + LEDD,
  ti_data, ~ visit_year + 1 | participant_id,
  weights = varIdent(form = ~ 1 | Cohort))

# time-interaction model
m_ti <- safe_lme(
  score ~ visit_year * Cohort + visit_year * baseline_score + visit_year * Age_onset +
    visit_year * Sex + visit_year * Race + LEDD,
  ti_data, ~ visit_year + 1 | participant_id,
  weights = varIdent(form = ~ 1 | Cohort))

cat("Current model (baseline adjustment):\n")
if (m_base$success) {
  tt_base <- summary(m_base$model)$tTable
  ir <- grep("visit_year:Cohort", rownames(tt_base))
  cat(sprintf("  MC1R x time: beta=%.3f, SE=%.3f, p=%.4f\n",
              tt_base[ir,"Value"], tt_base[ir,"Std.Error"], tt_base[ir,"p-value"]))
}

cat("\nTime-interaction model (covariates x time, except LEDD):\n")
if (m_ti$success) {
  tt_ti <- summary(m_ti$model)$tTable
  ir_ti <- grep("visit_year:Cohort", rownames(tt_ti))
  cat(sprintf("  MC1R x time: beta=%.3f, SE=%.3f, p=%.4f\n",
              tt_ti[ir_ti,"Value"], tt_ti[ir_ti,"Std.Error"], tt_ti[ir_ti,"p-value"]))

  cat("\nFull time-interaction coefficients:\n")
  ti_rows <- grep("visit_year", rownames(tt_ti))
  print(round(tt_ti[ti_rows, , drop = FALSE], 4))

  # LRT comparing nested models
  cat("\nLikelihood ratio test (base vs time-interaction):\n")
  lr <- anova(m_base$model, m_ti$model)
  print(lr)
} else {
  cat("  Time-interaction model failed to converge.\n")
  if (!is.null(m_ti$note)) cat("  convergence note:", m_ti$note, "\n")
}
```
```{r}
# PC-based ancestry sensitivity analyses
pcs <- read.csv(config$pcs)
pcs$Subject_ID <- pcs$sample_id

d_u3_pc <- d_u3 %>%
  left_join(pcs %>% select(Subject_ID, PC1, PC2, PC3, PC4, PC5,
                           PC6, PC7, PC8, PC9, PC10, EUR),
            by = c("participant_id" = "Subject_ID"))

pc_matched <- sum(!is.na(d_u3_pc$PC1) & d_u3_pc$Cohort %in% c("SPD", "MC1R PD"))
cat(sprintf("PC data merged: %d/%d sporadic participants matched\n",
            n_distinct(d_u3_pc$participant_id[!is.na(d_u3_pc$PC1) & 
                       d_u3_pc$Cohort %in% c("SPD", "MC1R PD")]),
            n_distinct(d_u3_pc$participant_id[d_u3_pc$Cohort %in% c("SPD", "MC1R PD")])))

# how many non-EUR in the sporadic cohort?
eur_flag <- d_u3_pc %>%
  filter(Cohort %in% c("SPD", "MC1R PD")) %>%
  distinct(participant_id, EUR, Cohort)
cat(sprintf("EUR: %d | non-EUR: %d | missing PC data: %d\n",
            sum(eur_flag$EUR == TRUE, na.rm = TRUE),
            sum(eur_flag$EUR == FALSE, na.rm = TRUE),
            sum(is.na(eur_flag$EUR))))

# EUR-restricted
d_eur <- d_u3_pc %>%
  filter(Cohort %in% c("SPD", "MC1R PD"), EUR == TRUE) %>%
  mutate(Cohort = factor(Cohort, levels = c("SPD", "MC1R PD"))) %>%
  filter(!is.na(score), !is.na(baseline_score), !is.na(Age_onset),
         !is.na(Sex), !is.na(Race), !is.na(LEDD))

m_eur <- safe_lme(
  score ~ visit_year * Cohort + baseline_score + Age_onset + Sex + Race + LEDD,
  d_eur, ~ visit_year + 1 | participant_id,
  weights = varIdent(form = ~ 1 | Cohort))

# PC-adjusted, full cohort
d_pc <- d_u3_pc %>%
  filter(Cohort %in% c("SPD", "MC1R PD"), !is.na(PC1)) %>%
  mutate(Cohort = factor(Cohort, levels = c("SPD", "MC1R PD"))) %>%
  filter(!is.na(score), !is.na(baseline_score), !is.na(Age_onset),
         !is.na(Sex), !is.na(Race), !is.na(LEDD))

m_pc <- safe_lme(
  score ~ visit_year * Cohort + baseline_score + Age_onset + Sex + Race + LEDD +
    PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10,
  d_pc, ~ visit_year + 1 | participant_id,
  weights = varIdent(form = ~ 1 | Cohort))

# compare against primary
pull_mc1r <- function(fit, label) {
  if (!fit$success) return(data.frame(Model = label, Beta = NA, SE = NA, P = NA, N = NA))
  tt <- summary(fit$model)$tTable
  ir <- grep("visit_year:Cohort", rownames(tt))
  data.frame(Model = label,
             N = n_distinct(fit$model$data$participant_id),
             Beta = round(tt[ir, "Value"], 3),
             SE = round(tt[ir, "Std.Error"], 3),
             P = round(tt[ir, "p-value"], 4))
}

pc_comparison <- bind_rows(
  pull_mc1r(m_base, "Primary (main effects)"),
  pull_mc1r(m_eur, "EUR-restricted"),
  pull_mc1r(m_pc, "PC1-10 adjusted")
)
print(pc_comparison)
```

## Stratified Analyses

```{r stratified}
run_strat <- function(strat_var, strat_levels, ledd = TRUE, min_n = MIN_N) {
  lapply(strat_levels, function(lev) {
    if (strat_var == "Sex") {
      sd <- d_u3 %>% filter(Sex == lev)
      grp <- paste0(lev, " MC1R PD")
    } else if (strat_var %in% c("Pen_inc", "Penetrance")) {
      mc1r_sub <- d_u3 %>%
        filter(Cohort == "MC1R PD", .data[[strat_var]] == lev) %>%
        mutate(Cohort = factor(paste0(lev, " MC1R PD")))
      sd <- bind_rows(d_u3 %>% filter(Cohort == "SPD"), mc1r_sub) %>%
        mutate(Cohort = factor(Cohort, levels = c("SPD", paste0(lev, " MC1R PD"))))
      grp <- paste0(lev, " MC1R PD")
    } else if (strat_var == "Genotype") {
      mc1r_sub <- d_u3 %>%
        filter(Cohort == "MC1R PD", Genotype == lev) %>%
        mutate(Cohort = factor(paste0(lev, " MC1R PD")))
      sd <- bind_rows(d_u3 %>% filter(Cohort == "SPD"), mc1r_sub) %>%
        mutate(Cohort = factor(Cohort, levels = c("SPD", paste0(lev, " MC1R PD"))))
      grp <- paste0(lev, " MC1R PD")
    } else if (grepl("_carrier$", strat_var)) {
      vn <- gsub("_carrier$", "", strat_var)
      mc1r_sub <- d_u3 %>%
        filter(Cohort == "MC1R PD", !!sym(strat_var) == TRUE) %>%
        mutate(Cohort = factor(paste0(vn, " MC1R PD")))
      sd <- bind_rows(d_u3 %>% filter(Cohort == "SPD"), mc1r_sub) %>%
        mutate(Cohort = factor(Cohort, levels = c("SPD", paste0(vn, " MC1R PD"))))
      grp <- paste0(vn, " MC1R PD")
    }

    nt <- n_distinct(sd$participant_id[sd$Cohort != "SPD"])
    if (nt < min_n)
      return(data.frame(Strat_var = strat_var, Stratum = as.character(lev),
                        Group = grp, N = nt, Slope = NA, Pct_diff = NA,
                        P_value = NA, Note = sprintf("N=%d", nt)))

    fstr <- if (ledd && "LEDD" %in% names(sd))
      "score ~ visit_year * Cohort + baseline_score + Age_onset + Sex + Race + LEDD"
    else "score ~ visit_year * Cohort + baseline_score + Age_onset + Sex + Race"
    if (strat_var == "Sex") fstr <- gsub(" \\+ Sex", "", fstr)

    sdc <- sd %>%
      filter(!is.na(score), !is.na(baseline_score), !is.na(Age_onset),
             !is.na(Cohort), !is.na(Sex), !is.na(Race))
    if (ledd && "LEDD" %in% names(sd)) sdc <- sdc %>% filter(!is.na(LEDD))

    mr <- safe_lme(as.formula(fstr), sdc, ~ visit_year + 1 | participant_id)
    if (mr$success) {
      sl <- extract_slopes(mr, "SPD", sdc)
      sl$Strat_var <- strat_var; sl$Stratum <- as.character(lev)
      sl %>% filter(Group != "SPD")
    } else {
      data.frame(Strat_var = strat_var, Stratum = as.character(lev),
                 Group = grp, N = nt, Slope = NA, Pct_diff = NA,
                 P_value = NA, Note = mr$error)
    }
  }) %>% bind_rows()
}

sl <- snl <- list()
sl$sex <- run_strat("Sex", c("Male","Female"), TRUE)
sl$pen <- run_strat("Pen_inc", c("r variant","R variant"), TRUE)
sl$pen_Rr <- run_strat("Penetrance", c("R+r variant"), TRUE)
sl$gen <- run_strat("Genotype", c("Heterozygote","Homozygote"), TRUE)

snl$sex <- run_strat("Sex", c("Male","Female"), FALSE)
snl$pen <- run_strat("Pen_inc", c("r variant","R variant"), FALSE)
snl$pen_Rr <- run_strat("Penetrance", c("R+r variant"), FALSE)
snl$gen <- run_strat("Genotype", c("Heterozygote","Homozygote"), FALSE)

for (vn in names(VARIANT_COLS)) {
  vc <- paste0(vn, "_carrier")
  if (vc %in% names(d_u3)) {
    sl[[vc]]  <- run_strat(vc, TRUE, TRUE, min_n = 3)
    snl[[vc]] <- run_strat(vc, TRUE, FALSE, min_n = 3)
  } else {
    ph <- data.frame(Strat_var = vc, Stratum = "TRUE", Group = paste0(vn, " MC1R PD"),
                     N = 0, Slope = NA, Pct_diff = NA, P_value = NA, Note = "Column not found")
    sl[[vc]] <- ph; snl[[vc]] <- ph
  }
}

u3_strat_df <- bind_rows(
  bind_rows(sl)  %>% mutate(LEDD_adjusted = "Yes"),
  bind_rows(snl) %>% mutate(LEDD_adjusted = "No"))

print(u3_strat_df %>% filter(LEDD_adjusted == "Yes") %>%
        select(Strat_var, Group, N, Slope, Pct_diff, P_value) %>%
        mutate(Slope = round(Slope, 2), P_value = sapply(P_value, fmt_p)))
```

## SCOPA-AUT

```{r scopa}
scopa_items <- paste0("SCAU", 1:21)
gi_items    <- paste0("SCAU", 1:7)
uri_items   <- paste0("SCAU", 8:13)
cv_items    <- paste0("SCAU", 14:16)

d_scopa <- load_ppmi_base(config$scopa_aut, VALID_IDS) %>%
  mutate(score = row_sum_threshold(., scopa_items, 18))

# subscores
d_scopa$scopa_gi  <- row_sum_threshold(d_scopa, gi_items, 5)
d_scopa$scopa_uri <- row_sum_threshold(d_scopa, uri_items, 4)
d_scopa$scopa_cv  <- row_sum_threshold(d_scopa, cv_items, 2)

d_scopa <- d_scopa %>%
  left_join(dem %>% select(Subject_ID, Cohort, Age_onset, Sex, Race),
            by = c("participant_id" = "Subject_ID")) %>%
  filter(!is.na(score))

scopa_results <- list()
for (outcome in c("score", "scopa_gi", "scopa_uri", "scopa_cv")) {
  for (cn in names(COMPARISONS)) {
    d <- d_scopa %>% mutate(score = .data[[outcome]])
    r <- run_progression(d, "score", COMPARISONS[[cn]], ledd = FALSE)
    if (r$success)
      scopa_results[[paste0(outcome, "_", cn)]] <- r$slopes %>%
        mutate(Outcome = outcome, Analysis_type = cn)
  }
}
scopa_df <- bind_rows(scopa_results)
```

## GDS-15

```{r gds}
gds_reverse <- c("GDSSATIS", "GDSGSPIR", "GDSHAPPY", "GDSALIVE", "GDSENRGY")
gds_direct  <- c("GDSDROPD", "GDSEMPTY", "GDSBORED", "GDSAFRAD", "GDSHLPLS",
                  "GDSHOME", "GDSMEMRY", "GDSWRTLS", "GDSHOPLS", "GDSBETER")

d_gds <- load_ppmi_base(config$gds, VALID_IDS)

# GDS: 5 reverse-scored (1-val) + 10 direct, require 13/15 non-NA
gds_all <- c(gds_reverse, gds_direct)
gds_mat <- as.matrix(d_gds[, gds_all, drop = FALSE])
storage.mode(gds_mat) <- "double"
gds_valid <- rowSums(!is.na(gds_mat))
rev_mat <- 1 - gds_mat[, gds_reverse, drop = FALSE]
dir_mat <- gds_mat[, gds_direct, drop = FALSE]
d_gds$score <- ifelse(gds_valid >= 13,
                       rowSums(rev_mat, na.rm = TRUE) + rowSums(dir_mat, na.rm = TRUE),
                       NA_real_)

d_gds <- d_gds %>%
  left_join(dem %>% select(Subject_ID, Cohort, Age_onset, Sex, Race),
            by = c("participant_id" = "Subject_ID")) %>%
  filter(!is.na(score))

gds_results <- list()
for (cn in names(COMPARISONS)) {
  comp <- COMPARISONS[[cn]]
  d <- d_gds
  r <- run_progression(d, "score", comp, ledd = FALSE)
  if (r$success)
    gds_results[[paste0(cn, "_noledd")]] <- r$slopes %>%
      mutate(Outcome = "GDS-15", Analysis_type = cn, LEDD_adjusted = "No")

  # borrow LEDD from nearest UPDRS III visit
  ledd_lk <- d_u3 %>% select(participant_id, visit_year, LEDD) %>%
    filter(!is.na(LEDD)) %>% distinct()
  dl <- d_gds %>%
    left_join(ledd_lk, by = "participant_id",
              suffix = c("", "_u"), relationship = "many-to-many") %>%
    mutate(dt = abs(visit_year - visit_year_u)) %>%
    group_by(participant_id, visit_year) %>%
    slice_min(dt, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    filter(dt < 0.5) %>%
    select(-visit_year_u, -dt)

  r2 <- run_progression(dl, "score", comp, ledd = TRUE)
  if (r2$success)
    gds_results[[paste0(cn, "_ledd")]] <- r2$slopes %>%
      mutate(Outcome = "GDS-15", Analysis_type = cn, LEDD_adjusted = "Yes")
}
gds_df <- bind_rows(gds_results)
```

## TD/PIGD Motor Subtype

```{r td-pigd}
# Stebbins 2013 classification
raw_p2 <- read_csv(config$updrs2, show_col_types = FALSE) %>%
  mutate(participant_id = paste0("PP-", PATNO)) %>%
  filter(participant_id %in% VALID_IDS, EVENT_ID == "BL")
raw_p3 <- read_csv(config$updrs3, show_col_types = FALSE) %>%
  mutate(participant_id = paste0("PP-", PATNO)) %>%
  filter(participant_id %in% VALID_IDS)

tremor_iii <- c("NP3PTRMR","NP3PTRML","NP3KTRMR","NP3KTRML",
                "NP3RTARU","NP3RTALU","NP3RTARL","NP3RTALL","NP3RTALJ","NP3RTCON")
pigd_iii <- c("NP3GAIT","NP3FRZGT","NP3PSTBL")

p2_bl <- raw_p2 %>%
  mutate(across(c(NP2TRMR, NP2WALK, NP2FREZ), as.numeric)) %>%
  group_by(participant_id) %>% slice(1) %>% ungroup() %>%
  select(participant_id, NP2TRMR, NP2WALK, NP2FREZ)

p3_bl <- raw_p3 %>%
  filter(EVENT_ID == "BL", PAG_NAME == "NUPDRS3",
         PDTRTMNT != "1" | is.na(PDTRTMNT)) %>%
  mutate(across(all_of(c(tremor_iii, pigd_iii)), as.numeric)) %>%
  group_by(participant_id) %>% slice(1) %>% ungroup() %>%
  select(participant_id, all_of(c(tremor_iii, pigd_iii)))

stebbins <- p2_bl %>%
  inner_join(p3_bl, by = "participant_id") %>%
  rowwise() %>%
  mutate(
    t_mean = mean(c(NP2TRMR, NP3PTRMR, NP3PTRML, NP3KTRMR, NP3KTRML,
                     NP3RTARU, NP3RTALU, NP3RTARL, NP3RTALL, NP3RTALJ,
                     NP3RTCON), na.rm = TRUE),
    p_mean = mean(c(NP2WALK, NP2FREZ, NP3GAIT, NP3FRZGT, NP3PSTBL), na.rm = TRUE),
    ratio  = if_else(p_mean > 0, t_mean / p_mean, NA_real_),
    subtype = case_when(
      t_mean == 0 & p_mean == 0 ~ "Indeterminate",
      p_mean == 0 & t_mean > 0  ~ "TD",
      ratio >= 1.15 ~ "TD", ratio <= 0.90 ~ "PIGD",
      TRUE ~ "Indeterminate")
  ) %>% ungroup() %>%
  select(participant_id, subtype, t_mean, p_mean, ratio)

d_u3_sub <- d_u3 %>%
  left_join(stebbins %>% select(participant_id, subtype), by = "participant_id") %>%
  filter(!is.na(subtype), subtype != "Indeterminate")

subtype_results <- list()
comp <- COMPARISONS[["primary"]]
for (st in c("TD", "PIGD")) {
  d <- d_u3_sub %>% filter(subtype == st)
  for (ledd in c(TRUE, FALSE)) {
    r <- run_progression(d, "score", comp, ledd = ledd)
    if (r$success)
      subtype_results[[paste0(st, "_", ledd)]] <- r$slopes %>%
        mutate(Motor_subtype = st, Model = if (ledd) "LEDD-adjusted" else "No LEDD")
  }
}

# formal 3-way interaction
int_data <- d_u3_sub %>%
  filter(Cohort %in% c(comp$ref, comp$test)) %>%
  mutate(Cohort = factor(Cohort, levels = c(comp$ref, comp$test)),
         subtype = factor(subtype, levels = c("TD", "PIGD"))) %>%
  group_by(participant_id) %>%
  mutate(baseline_score = first(score, order_by = visit_year)) %>%
  ungroup() %>%
  filter(!is.na(baseline_score), !is.na(Age_onset), !is.na(Sex), !is.na(Race))

int_fit <- safe_lme(score ~ visit_year * Cohort * subtype + baseline_score + Age_onset + Sex + Race,
                    int_data, ~ visit_year + 1 | participant_id)
if (int_fit$success) print(round(summary(int_fit$model)$tTable, 4))

subtype_df <- bind_rows(subtype_results)
```

## Dose-Response and Penetrance Trends

```{r dose-trends}
# allele dose: SPD(0) -> Het(1) -> Hom(2)
d_dose <- d_u3 %>%
  filter(Cohort == "SPD" | (Cohort == "MC1R PD" & Genotype != "Non-carrier")) %>%
  mutate(dose = case_when(Cohort == "SPD" ~ 0L,
                          Genotype == "Heterozygote" ~ 1L,
                          Genotype == "Homozygote" ~ 2L)) %>%
  group_by(participant_id) %>%
  mutate(baseline_score = first(score, order_by = visit_year)) %>%
  ungroup() %>%
  filter(!is.na(dose), !is.na(baseline_score),
         !is.na(Age_onset), !is.na(Sex), !is.na(Race), !is.na(LEDD))

dose_fit <- safe_lme(score ~ visit_year * dose + baseline_score + Age_onset + Sex + Race + LEDD,
                      d_dose, ~ visit_year + 1 | participant_id)

# categorical version
d_dose_cat <- d_dose %>%
  mutate(dose_cat = factor(dose, levels = 0:2, labels = c("SPD","Het","Hom")))
dose_cat_fit <- safe_lme(score ~ visit_year * dose_cat + baseline_score + Age_onset + Sex + Race + LEDD,
                          d_dose_cat, ~ visit_year + 1 | participant_id)

# penetrance: SPD(0) -> r(1) -> R(2)
d_pen <- d_u3 %>%
  filter(Cohort == "SPD" | (Cohort == "MC1R PD" & Pen_inc != "Unknown")) %>%
  mutate(pen = case_when(Cohort == "SPD" ~ 0L,
                         Pen_inc == "r variant" ~ 1L,
                         Pen_inc == "R variant" ~ 2L)) %>%
  group_by(participant_id) %>%
  mutate(baseline_score = first(score, order_by = visit_year)) %>%
  ungroup() %>%
  filter(!is.na(pen), !is.na(baseline_score),
         !is.na(Age_onset), !is.na(Sex), !is.na(Race), !is.na(LEDD))

pen_fit <- safe_lme(score ~ visit_year * pen + baseline_score + Age_onset + Sex + Race + LEDD,
                     d_pen, ~ visit_year + 1 | participant_id)

d_pen_cat <- d_pen %>%
  mutate(pen_cat = factor(pen, levels = 0:2, labels = c("SPD","r-variant","R-variant")))
pen_cat_fit <- safe_lme(score ~ visit_year * pen_cat + baseline_score + Age_onset + Sex + Race + LEDD,
                         d_pen_cat, ~ visit_year + 1 | participant_id)

# collect trend results
trend_results <- data.frame(Test = character(), Model = character(),
                             Beta = numeric(), SE = numeric(), P_value = character())
for (ti in list(
  list("Dose", "Linear",     dose_fit,     "visit_year:dose"),
  list("Dose", "Cat_Het",    dose_cat_fit, "visit_year:dose_catHet"),
  list("Dose", "Cat_Hom",    dose_cat_fit, "visit_year:dose_catHom"),
  list("Pen",  "Linear",     pen_fit,      "visit_year:pen"),
  list("Pen",  "Cat_r",      pen_cat_fit,  "visit_year:pen_catr-variant"),
  list("Pen",  "Cat_R",      pen_cat_fit,  "visit_year:pen_catR-variant")
)) {
  if (ti[[3]]$success) {
    t <- pull_term(ti[[3]]$model, ti[[4]])
    if (!is.null(t))
      trend_results <- rbind(trend_results, data.frame(
        Test = ti[[1]], Model = ti[[2]],
        Beta = round(t$b, 4), SE = round(t$se, 4), P_value = fmt_p(t$p)))
  }
}
print(trend_results)
```

## Nonlinear Time

```{r quadratic}
d_quad <- d_u3 %>%
  filter(Cohort %in% c("SPD", "MC1R PD")) %>%
  mutate(Cohort = factor(Cohort, levels = c("SPD", "MC1R PD")),
         vy2 = visit_year^2) %>%
  group_by(participant_id) %>%
  mutate(baseline_score = first(score, order_by = visit_year)) %>%
  ungroup() %>%
  filter(!is.na(score), !is.na(baseline_score),
         !is.na(Age_onset), !is.na(Sex), !is.na(Race), !is.na(LEDD))

lin_fit <- safe_lme(score ~ visit_year * Cohort + baseline_score + Age_onset + Sex + Race + LEDD,
                     d_quad, ~ visit_year + 1 | participant_id)
quad_fit <- safe_lme(score ~ (visit_year + vy2) * Cohort + baseline_score + Age_onset + Sex + Race + LEDD,
                      d_quad, ~ visit_year + 1 | participant_id)

if (quad_fit$success) {
  print(round(summary(quad_fit$model)$tTable, 4))
  if (lin_fit$success)
    cat(sprintf("Linear AIC=%.1f BIC=%.1f | Quad AIC=%.1f BIC=%.1f\n",
                AIC(lin_fit$model), BIC(lin_fit$model),
                AIC(quad_fit$model), BIC(quad_fit$model)))
}
```

## Fast Progressor Enrichment

```{r fast-prog}
fp_data <- d_u3 %>%
  filter(Cohort %in% c("SPD", "MC1R PD")) %>%
  mutate(Cohort = factor(Cohort, levels = c("SPD", "MC1R PD"))) %>%
  group_by(participant_id) %>%
  mutate(baseline_score = first(score, order_by = visit_year)) %>%
  ungroup() %>%
  filter(!is.na(score), !is.na(baseline_score),
         !is.na(Age_onset), !is.na(Sex), !is.na(Race), !is.na(LEDD))

# no Cohort interaction -> unbiased individual slopes
fp_fit <- safe_lme(score ~ visit_year + baseline_score + Age_onset + Sex + Race + LEDD,
                    fp_data, ~ visit_year + 1 | participant_id)

fp_df <- fp_trend <- data.frame()
if (fp_fit$success) {
  pc <- coef(fp_fit$model)
  islopes <- data.frame(
    pid = rownames(pc), slope = pc[,"visit_year"])
  cmap <- fp_data %>% distinct(participant_id, Cohort)
  islopes$Cohort <- cmap$Cohort[match(islopes$pid, cmap$participant_id)]
  islopes <- islopes[!is.na(islopes$Cohort), ]

  q75 <- quantile(islopes$slope, 0.75)
  q90 <- quantile(islopes$slope, 0.90)
  med <- median(islopes$slope)

  fp_results <- list()
  for (info in list(
    list("Above median", med), list("Top quartile", q75), list("Top decile", q90)
  )) {
    lbl <- info[[1]]; cut <- info[[2]]
    fast_mc <- sum(islopes$slope >= cut & islopes$Cohort == "MC1R PD")
    tot_mc  <- sum(islopes$Cohort == "MC1R PD")
    fast_sp <- sum(islopes$slope >= cut & islopes$Cohort == "SPD")
    tot_sp  <- sum(islopes$Cohort == "SPD")

    tab <- matrix(c(fast_mc, tot_mc - fast_mc, fast_sp, tot_sp - fast_sp), nrow = 2, byrow = TRUE)
    ft <- fisher.test(tab)

    fp_results[[lbl]] <- data.frame(
      Threshold = lbl, Cutoff = round(cut, 2),
      MC1R_n = fast_mc, MC1R_total = tot_mc, MC1R_pct = round(100*fast_mc/tot_mc, 1),
      SPD_n = fast_sp, SPD_total = tot_sp, SPD_pct = round(100*fast_sp/tot_sp, 1),
      OR = round(ft$estimate, 2),
      OR_lower = round(ft$conf.int[1], 2), OR_upper = round(ft$conf.int[2], 2),
      P_fisher = fmt_p(ft$p.value))
  }
  fp_df <- bind_rows(fp_results)
  print(fp_df)

  # ordinal trend test
  islopes$prog_cat <- case_when(
    islopes$slope >= q90 ~ 3L, islopes$slope >= q75 ~ 2L,
    islopes$slope >= med ~ 1L, TRUE ~ 0L)
  islopes$is_mc1r <- as.integer(islopes$Cohort == "MC1R PD")

  tfit <- glm(is_mc1r ~ prog_cat, data = islopes, family = binomial)
  tc <- summary(tfit)$coefficients["prog_cat", ]
  fp_trend <- data.frame(
    Test = "Trend (ordinal logistic)",
    OR_per_category = round(exp(tc["Estimate"]), 2),
    CI_lower = round(exp(tc["Estimate"] - 1.96*tc["Std. Error"]), 2),
    CI_upper = round(exp(tc["Estimate"] + 1.96*tc["Std. Error"]), 2),
    P_trend = fmt_p(tc["Pr(>|z|)"]))
  print(fp_trend)
}
```

## Phenoconversion (Fine-Gray)

```{r phenoconversion}
d_prod <- read_excel(config$data, sheet = "Prodromal") %>%
  mutate(Cohort = trimws(Cohort),
         MC1R_carrier = as.integer(Cohort == "MC1R"),
         status = as.integer(Phenoconverter),
         time_years = as.numeric(difftime(
           if_else(status >= 1, `Date of Phenoconversion`, `Analysis End`),
           `Analysis Start`, units = "days")) / 365.25,
         Age = `Age at baseline`) %>%
  filter(!is.na(time_years), time_years > 0, !is.na(Age), !is.na(Sex))

d_prod$fstatus <- factor(d_prod$status, levels = 0:2, labels = c("cens", "PD", "DLB"))

cat(sprintf("Prodromal cohort: %d participants (%d PD, %d DLB/MSA, %d censored)\n",
            nrow(d_prod), sum(d_prod$status == 1), sum(d_prod$status == 2),
            sum(d_prod$status == 0)))
cat(sprintf("  MC1R carriers: %d (%d PD, %d DLB/MSA)\n",
            sum(d_prod$MC1R_carrier),
            sum(d_prod$MC1R_carrier & d_prod$status == 1),
            sum(d_prod$MC1R_carrier & d_prod$status == 2)))
cat(sprintf("  Non-carriers: %d (%d PD, %d DLB/MSA)\n",
            sum(!d_prod$MC1R_carrier),
            sum(!d_prod$MC1R_carrier & d_prod$status == 1),
            sum(!d_prod$MC1R_carrier & d_prod$status == 2)))

# Fine-Gray subdistribution hazard for PD with DLB/MSA as competing event
fg_data <- finegray(Surv(time_years, fstatus) ~ ., data = d_prod, etype = "PD")
fg_fit <- coxph(Surv(fgstart, fgstop, fgstatus) ~ MC1R_carrier + Age + Sex,
                data = fg_data, weights = fgwt)
fg_sum <- summary(fg_fit)

cat("\nFine-Gray subdistribution hazard (PD):\n")
print(fg_sum)

fg_hr <- exp(coef(fg_fit))["MC1R_carrier"]
fg_ci <- exp(confint(fg_fit))["MC1R_carrier", ]
cat(sprintf("\nMC1R carrier sHR: %.2f [95%% CI: %.2f, %.2f], p=%.4f\n",
            fg_hr, fg_ci[1], fg_ci[2],
            fg_sum$coefficients["MC1R_carrier", "Pr(>|z|)"]))

# CIF for figure
cif_fit <- survfit(Surv(time_years, fstatus) ~ MC1R_carrier, data = d_prod)
cat("\nCumulative incidence summary:\n")
print(cif_fit)
```

```{r sensitivity-ti-subgroups}
# covariate Ă— time interaction stability for dose, penetrance, and variant subgroups
ti_subgroup <- function(data, strat_col, strat_val, label, use_ledd = TRUE) {
  mc1r_sub <- data %>%
    filter(Cohort == "MC1R PD", .data[[strat_col]] == strat_val)
  ref_sub <- data %>% filter(Cohort == "SPD")
  d <- bind_rows(mc1r_sub, ref_sub) %>%
    mutate(MC1R_bin = as.integer(Cohort == "MC1R PD")) %>%
    group_by(participant_id) %>%
    mutate(baseline_score = first(score, order_by = visit_year)) %>%
    ungroup()

  req <- c("score", "visit_year", "baseline_score", "Age_onset", "Sex", "Race", "participant_id")
  if (use_ledd) req <- c(req, "LEDD")
  d <- d %>% filter(if_all(all_of(req), ~ !is.na(.)))

  nt <- n_distinct(d$participant_id[d$MC1R_bin == 1])
  nr <- n_distinct(d$participant_id[d$MC1R_bin == 0])
  if (nt < 10 || nr < 10) return(NULL)

  covs_base <- "baseline_score + Age_onset + Sex + Race"
  covs_ti   <- "visit_year * baseline_score + visit_year * Age_onset + visit_year * Sex + visit_year * Race"
  if (use_ledd) {
    covs_base <- paste(covs_base, "+ LEDD")
    covs_ti   <- paste(covs_ti, "+ LEDD")
  }

  fml_base <- as.formula(paste("score ~ visit_year * MC1R_bin +", covs_base))
  fml_ti   <- as.formula(paste("score ~ visit_year * MC1R_bin +", covs_ti))

  fit_base <- safe_lme(fml_base, d, ~ visit_year + 1 | participant_id)
  fit_ti   <- safe_lme(fml_ti, d, ~ visit_year + 1 | participant_id)

  pull_mc1r <- function(fit) {
    if (!fit$success) return(c(Beta = NA, SE = NA, P = NA))
    tt <- summary(fit$model)$tTable
    ir <- grep("visit_year:MC1R_bin|MC1R_bin:visit_year", rownames(tt))
    if (length(ir) == 0) return(c(Beta = NA, SE = NA, P = NA))
    c(Beta = tt[ir, "Value"], SE = tt[ir, "Std.Error"], P = tt[ir, "p-value"])
  }

  b <- pull_mc1r(fit_base); t <- pull_mc1r(fit_ti)

  data.frame(
    Subgroup = label, N_test = nt, N_ref = nr,
    Beta_base = round(b["Beta"], 4), SE_base = round(b["SE"], 4), P_base = b["P"],
    Beta_ti   = round(t["Beta"], 4), SE_ti   = round(t["SE"], 4), P_ti   = t["P"],
    Pct_change = if (!is.na(b["Beta"]) && !is.na(t["Beta"]) && b["Beta"] != 0)
                   round(100 * (t["Beta"] - b["Beta"]) / abs(b["Beta"]), 1) else NA_real_
  )
}

ti_sub <- bind_rows(
  # dose
  ti_subgroup(d_u3, "Genotype", "Heterozygote", "Heterozygote"),
  ti_subgroup(d_u3, "Genotype", "Homozygote", "Homozygote"),
  # penetrance
  ti_subgroup(d_u3, "Pen_inc", "r variant", "r variant"),
  ti_subgroup(d_u3, "Pen_inc", "R variant", "R variant"),
  # individual variants
  {
    var_rows <- lapply(names(VARIANT_COLS), function(vn) {
      carrier_col <- paste0(vn, "_carrier")
      if (!carrier_col %in% names(d_u3)) return(NULL)
      d_v <- d_u3 %>%
        mutate(is_carrier = .data[[carrier_col]]) %>%
        filter(Cohort == "SPD" | (Cohort == "MC1R PD" & is_carrier))
      n_carrier <- n_distinct(d_v$participant_id[d_v$Cohort == "MC1R PD"])
      if (n_carrier < 10) return(NULL)
      # temporarily relabel so ti_subgroup logic works
      d_v$variant_flag <- ifelse(d_v$Cohort == "MC1R PD" & d_v$is_carrier, "carrier", "non")
      ti_subgroup(d_v, "variant_flag", "carrier", vn)
    })
    bind_rows(var_rows)
  }
)

cat("MC1R x time stability across subgroups:\n")
print(ti_sub %>% select(Subgroup, N_test, N_ref, Beta_base, P_base, Beta_ti, P_ti, Pct_change))
```
```{r sensitivity-ti-all}
# covariate Ă— time interactions across all outcomes
# compare MC1R Ă— time coefficient: base model vs time-interaction model
ti_compare <- function(data, outcome, bl_col, label, use_ledd = FALSE) {
  d <- data %>%
    filter(Cohort %in% c("SPD", "MC1R PD")) %>%
    mutate(Cohort = factor(Cohort, levels = c("SPD", "MC1R PD")))

  if (is.null(bl_col)) {
    d <- d %>%
      group_by(participant_id) %>%
      mutate(bl = first(!!sym(outcome), order_by = visit_year)) %>%
      ungroup()
    bl_col <- "bl"
  }

  req <- c(outcome, "visit_year", bl_col, "Age_onset", "Sex", "Race", "Cohort", "participant_id")
  if (use_ledd) req <- c(req, "LEDD")
  d <- d %>% filter(if_all(all_of(req), ~ !is.na(.)))
  if (n_distinct(d$participant_id) < 50) return(NULL)

  covs_base <- paste(bl_col, "+ Age_onset + Sex + Race")
  covs_ti   <- paste("visit_year *", bl_col, "+ visit_year * Age_onset + visit_year * Sex + visit_year * Race")
  if (use_ledd) {
    covs_base <- paste(covs_base, "+ LEDD")
    covs_ti   <- paste(covs_ti, "+ LEDD")   # LEDD time-varying, no interaction
  }

  fml_base <- as.formula(sprintf("%s ~ visit_year * Cohort + %s", outcome, covs_base))
  fml_ti   <- as.formula(sprintf("%s ~ visit_year * Cohort + %s", outcome, covs_ti))

  fit_base <- safe_lme(fml_base, d, ~ visit_year + 1 | participant_id,
                        weights = varIdent(form = ~ 1 | Cohort))
  fit_ti   <- safe_lme(fml_ti, d, ~ visit_year + 1 | participant_id,
                        weights = varIdent(form = ~ 1 | Cohort))

  extract_mc1r <- function(fit, tag) {
    if (!fit$success) return(data.frame(Model = tag, Beta = NA, SE = NA, P = NA))
    tt <- summary(fit$model)$tTable
    ir <- grep("visit_year:Cohort", rownames(tt))
    if (length(ir) == 0) return(data.frame(Model = tag, Beta = NA, SE = NA, P = NA))
    data.frame(Model = tag, Beta = tt[ir, "Value"], SE = tt[ir, "Std.Error"], P = tt[ir, "p-value"])
  }

  base_r <- extract_mc1r(fit_base, "Base")
  ti_r   <- extract_mc1r(fit_ti, "Cov x Time")

  data.frame(
    Outcome  = label,
    N        = n_distinct(d$participant_id),
    LEDD     = use_ledd,
    Beta_base = round(base_r$Beta, 4), SE_base = round(base_r$SE, 4), P_base = base_r$P,
    Beta_ti   = round(ti_r$Beta, 4),   SE_ti   = round(ti_r$SE, 4),   P_ti   = ti_r$P,
    Pct_change = if (!is.na(base_r$Beta) && !is.na(ti_r$Beta) && base_r$Beta != 0)
                   round(100 * (ti_r$Beta - base_r$Beta) / abs(base_r$Beta), 1) else NA_real_
  )
}

# subscores need their own baseline column
ti_subscore <- function(data, sub, label, use_ledd = FALSE) {
  bl <- paste0("bl_", sub)
  d <- data %>%
    filter(Cohort %in% c("SPD", "MC1R PD"), !is.na(.data[[sub]])) %>%
    mutate(Cohort = factor(Cohort, levels = c("SPD", "MC1R PD"))) %>%
    group_by(participant_id) %>%
    mutate(!!bl := first(.data[[sub]], order_by = visit_year)) %>%
    ungroup()

  req <- c(sub, "visit_year", bl, "Age_onset", "Sex", "Race", "Cohort", "participant_id")
  if (use_ledd) req <- c(req, "LEDD")
  d <- d %>% filter(if_all(all_of(req), ~ !is.na(.)))
  if (n_distinct(d$participant_id) < 50) return(NULL)

  covs_base <- paste(bl, "+ Age_onset + Sex + Race")
  covs_ti   <- paste("visit_year *", bl, "+ visit_year * Age_onset + visit_year * Sex + visit_year * Race")
  if (use_ledd) {
    covs_base <- paste(covs_base, "+ LEDD")
    covs_ti   <- paste(covs_ti, "+ LEDD")
  }

  fml_base <- as.formula(paste0(sub, " ~ visit_year * Cohort + ", covs_base))
  fml_ti   <- as.formula(paste0(sub, " ~ visit_year * Cohort + ", covs_ti))

  fit_base <- safe_lme(fml_base, d, ~ visit_year + 1 | participant_id)
  fit_ti   <- safe_lme(fml_ti, d, ~ visit_year + 1 | participant_id)

  extract_mc1r <- function(fit, tag) {
    if (!fit$success) return(data.frame(Model = tag, Beta = NA, SE = NA, P = NA))
    tt <- summary(fit$model)$tTable
    ir <- grep("visit_year:Cohort", rownames(tt))
    if (length(ir) == 0) return(data.frame(Model = tag, Beta = NA, SE = NA, P = NA))
    data.frame(Model = tag, Beta = tt[ir, "Value"], SE = tt[ir, "Std.Error"], P = tt[ir, "p-value"])
  }

  base_r <- extract_mc1r(fit_base, "Base")
  ti_r   <- extract_mc1r(fit_ti, "Cov x Time")

  data.frame(
    Outcome  = label,
    N        = n_distinct(d$participant_id),
    LEDD     = use_ledd,
    Beta_base = round(base_r$Beta, 4), SE_base = round(base_r$SE, 4), P_base = base_r$P,
    Beta_ti   = round(ti_r$Beta, 4),   SE_ti   = round(ti_r$SE, 4),   P_ti   = ti_r$P,
    Pct_change = if (!is.na(base_r$Beta) && !is.na(ti_r$Beta) && base_r$Beta != 0)
                   round(100 * (ti_r$Beta - base_r$Beta) / abs(base_r$Beta), 1) else NA_real_
  )
}

ti_all <- bind_rows(
  # primary
  ti_compare(d_u3, "score", "baseline_score", "UPDRS III OFF", use_ledd = TRUE),
  ti_compare(d_u3_on, "score", "baseline_score", "UPDRS III ON", use_ledd = TRUE),

  # OFF subscores
  ti_subscore(d_u3, "bradykinesia_score", "Bradykinesia", use_ledd = TRUE),
  ti_subscore(d_u3, "rigidity_score", "Rigidity", use_ledd = TRUE),
  ti_subscore(d_u3, "tremor_score", "Tremor", use_ledd = TRUE),
  ti_subscore(d_u3, "axial_score", "Axial", use_ledd = TRUE),
  ti_subscore(d_u3, "bulbar_score", "Bulbar", use_ledd = TRUE),

  # secondary outcomes
  ti_compare(d_u1, "score", "baseline_score", "UPDRS I"),
  ti_compare(d_u2, "score", NULL, "UPDRS II"),
  ti_compare(d_u4, "score", NULL, "UPDRS IV"),
  ti_compare(d_moca, "score", "baseline_score", "MoCA"),
  ti_compare(d_dat_c, "score", "baseline_score", "DAT Caudate"),
  ti_compare(d_dat_p, "score", "baseline_score", "DAT Putamen")
)

# SCOPA/GDS/ESS if they exist
if (exists("d_scopa") && nrow(d_scopa) > 0)
  ti_all <- bind_rows(ti_all, ti_compare(d_scopa, "score", NULL, "SCOPA-AUT"))
if (exists("d_gds") && nrow(d_gds) > 0)
  ti_all <- bind_rows(ti_all, ti_compare(d_gds, "score", NULL, "GDS"))
if (exists("d_ess") && nrow(d_ess) > 0)
  ti_all <- bind_rows(ti_all, ti_compare(d_ess, "score", NULL, "ESS"))

ti_all <- ti_all %>%
  mutate(across(matches("P_"), ~ round(., 4)),
         across(matches("Beta_|SE_"), ~ round(., 4)))

cat("MC1R x time: base model vs covariate x time interaction:\n")
print(ti_all %>% select(Outcome, N, Beta_base, P_base, Beta_ti, P_ti, Pct_change))
```

```{r phenoconversion-figure, fig.width=8, fig.height=7}
if (!is.null(fg_fit)) {
  cols <- c("0" = "#2166AC", "1" = "#B2182B")

  # CIF from competing risk survfit
  cif_dat <- survfit(Surv(time_years, fstatus) ~ MC1R_carrier, data = d_prod)

  # extract PD cumulative incidence per group
  nc_n <- cif_dat$strata[1]
  mc_n <- cif_dat$strata[2]

  cif_df <- bind_rows(
    data.frame(time = 0, cuminc = 0, lower = 0, upper = 0, group = "0"),
    data.frame(time = 0, cuminc = 0, lower = 0, upper = 0, group = "1"),
    data.frame(
      time = cif_dat$time[1:nc_n],
      cuminc = cif_dat$pstate[1:nc_n, 2],
      lower = cif_dat$lower[1:nc_n, 2],
      upper = cif_dat$upper[1:nc_n, 2],
      group = "0"),
    data.frame(
      time = cif_dat$time[(nc_n+1):(nc_n+mc_n)],
      cuminc = cif_dat$pstate[(nc_n+1):(nc_n+mc_n), 2],
      lower = cif_dat$lower[(nc_n+1):(nc_n+mc_n), 2],
      upper = cif_dat$upper[(nc_n+1):(nc_n+mc_n), 2],
      group = "1")
  )

  fg_hr_val <- exp(coef(fg_fit))["MC1R_carrier"]
  fg_ci_vals <- exp(confint(fg_fit))["MC1R_carrier", ]
  fg_p <- summary(fg_fit)$coefficients["MC1R_carrier", "Pr(>|z|)"]
  stat_label <- sprintf("PD sHR = %.2f (95%% CI %.2f\u2013%.2f)\np = %s",
                        fg_hr_val, fg_ci_vals[1], fg_ci_vals[2], fmt_p(fg_p))

  risk_times <- seq(0, 11, 1)
  risk_tbl <- expand.grid(group = c("0", "1"), time = risk_times,
                          stringsAsFactors = FALSE) %>%
    rowwise() %>%
    mutate(n_risk = sum(d_prod$time_years >= time &
                        d_prod$MC1R_carrier == as.integer(group))) %>%
    ungroup()

  p_cif <- ggplot() +
    geom_stepribbon(data = cif_df, aes(x = time, ymin = lower, ymax = upper, fill = group),
                    alpha = 0.15, show.legend = FALSE) +
    geom_step(data = cif_df, aes(x = time, y = cuminc, color = group), linewidth = 1.5) +
    annotate("richtext", x = 0.15, y = 0.75,
             label = stat_label, hjust = 0, vjust = 1,
             size = 4, family = "sans", fill = "white",
             label.size = 0.5, label.r = unit(0.15, "lines")) +
    scale_color_manual(values = cols,
                       labels = c("0" = "Non-carriers",
                                  "1" = "<i>MC1R</i> carriers")) +
    scale_fill_manual(values = cols, guide = "none") +
    scale_x_continuous(breaks = seq(0, 12, 1), limits = c(0, 12),
                       expand = expansion(mult = c(0.02, 0.02))) +
    scale_y_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 0.85),
                       labels = scales::number_format(accuracy = 0.1),
                       expand = expansion(mult = c(0, 0.02))) +
    labs(x = "Years since Baseline",
         y = "Cumulative Incidence of PD") +
    theme_classic(base_size = 14) +
    theme(legend.position = c(0.17, 0.95),
          legend.background = element_rect(fill = "white", color = "gray80"),
          legend.title = element_blank(),
          legend.text = element_markdown(),
          legend.margin = margin(3, 6, 3, 6),
          axis.title.x = element_text(face = "bold"),
          axis.title.y = element_text(face = "bold", vjust = -14),
          axis.text = element_text(color = "black"),
          panel.grid.major = element_line(color = "gray90", linewidth = 0.3),
          plot.margin = margin(5, 10, -5, 10))

  risk_tbl$group <- factor(risk_tbl$group, levels = c("1", "0"))
  p_risk <- ggplot(risk_tbl, aes(x = time, y = group, label = n_risk)) +
    geom_text(aes(color = group), size = 3.5, fontface = "plain") +
    scale_color_manual(values = cols, guide = "none") +
    scale_x_continuous(breaks = seq(0, 12, 1), limits = c(0, 12),
                       expand = expansion(mult = c(0.01, 0.01))) +
    scale_y_discrete(labels = c("1" = expression(paste(italic("MC1R"), " carriers")),
                                "0" = "Non-carriers")) +
    labs(x = NULL, y = NULL) +
    theme_void(base_size = 11) +
    theme(axis.text.y = element_text(hjust = 1, face = "bold", color = "gray30",
                                     margin = margin(r = 10)),
          axis.text.x = element_blank(),
          plot.title = element_text(size = 10, face = "bold", hjust = -0.15,
                                    color = "gray30", margin = margin(b = 3)),
          plot.margin = margin(-5, 10, 5, 60)) +
    ggtitle("No. at risk")

  fig_cif <- p_cif / p_risk + plot_layout(heights = c(8, 1))

  print(fig_cif)
  ggsave(file.path(config$out_dir, "Figure_Phenoconversion_CIF.png"),
         fig_cif, width = 8, height = 6, dpi = 600)
  ggsave(file.path(config$out_dir, "Figure_Phenoconversion_CIF.pdf"),
         fig_cif, width = 8, height = 6, device = cairo_pdf)
}
```

## ESS

```{r ess}
d_ess <- load_ppmi_base(config$ess, VALID_IDS) %>%
  mutate(score = row_sum_threshold(., paste0("ESS", 1:8), 7))

d_ess <- d_ess %>%
  left_join(dem %>% select(Subject_ID, Cohort, Age_onset, Sex, Race),
            by = c("participant_id" = "Subject_ID")) %>%
  filter(!is.na(score))

ess_results <- list()
for (cn in names(COMPARISONS)) {
  r <- run_progression(d_ess, "score", COMPARISONS[[cn]], ledd = FALSE)
  if (r$success)
    ess_results[[cn]] <- r$slopes %>% mutate(Outcome = "ESS", Analysis_type = cn)
}
ess_df <- bind_rows(ess_results)
```

## Figure

```{r figure, fig.width=8, fig.height=6}
fdc <- d_u3 %>%
  filter(Cohort %in% c("SPD", "MC1R PD")) %>%
  mutate(Cohort = factor(Cohort, levels = c("SPD", "MC1R PD"))) %>%
  filter(!is.na(score), !is.na(baseline_score), !is.na(Age_onset),
         !is.na(Sex), !is.na(Race), !is.na(LEDD)) %>%
  droplevels()

fig_covars <- c("baseline_score", "Age_onset", "LEDD")
for (v in c("Sex", "Race"))
  if (length(unique(fdc[[v]])) >= 2) fig_covars <- c(fig_covars, v)
fig_fml <- as.formula(paste("score ~ visit_year * Cohort +", paste(fig_covars, collapse = " + ")))

fig_mod <- tryCatch({
  lme(fig_fml, data = fdc, random = ~ visit_year + 1 | participant_id,
      weights = varIdent(form = ~ 1 | Cohort),
      control = lmeControl(opt = "optim", maxIter = 200, msMaxIter = 200), na.action = na.omit)
}, error = function(e) NULL)

if (!is.null(fig_mod)) {
  fe <- fixef(fig_mod)
  yrs <- seq(0, max(fdc$visit_year, na.rm = TRUE), by = 0.5)
  cov_val <- fe["baseline_score"] * mean(fdc$baseline_score, na.rm = TRUE) +
    fe["Age_onset"] * mean(fdc$Age_onset, na.rm = TRUE) +
    fe["LEDD"] * mean(fdc$LEDD, na.rm = TRUE)

  pg <- data.frame(
    visit_year = rep(yrs, 2),
    Cohort = factor(rep(c("SPD","MC1R PD"), each = length(yrs)), levels = c("SPD","MC1R PD")))
  pg$predicted <- ifelse(
    pg$Cohort == "SPD",
    fe["(Intercept)"] + fe["visit_year"] * pg$visit_year + cov_val,
    fe["(Intercept)"] + fe["CohortMC1R PD"] +
      (fe["visit_year"] + fe["visit_year:CohortMC1R PD"]) * pg$visit_year + cov_val)

  vm <- vcov(fig_mod)
  pg$se <- NA
  for (i in 1:nrow(pg)) {
    t <- pg$visit_year[i]
    if (pg$Cohort[i] == "SPD") {
      vp <- vm["(Intercept)","(Intercept)"] + t^2*vm["visit_year","visit_year"] +
        2*t*vm["(Intercept)","visit_year"]
    } else {
      vp <- vm["(Intercept)","(Intercept)"] + vm["CohortMC1R PD","CohortMC1R PD"] +
        t^2*vm["visit_year","visit_year"] + t^2*vm["visit_year:CohortMC1R PD","visit_year:CohortMC1R PD"] +
        2*vm["(Intercept)","CohortMC1R PD"] + 2*t*vm["(Intercept)","visit_year"] +
        2*t*vm["(Intercept)","visit_year:CohortMC1R PD"] + 2*t*vm["CohortMC1R PD","visit_year"] +
        2*t*vm["CohortMC1R PD","visit_year:CohortMC1R PD"] + 2*t^2*vm["visit_year","visit_year:CohortMC1R PD"]
    }
    pg$se[i] <- sqrt(max(vp, 0))
  }
  pg$ci_lo <- pg$predicted - 1.96 * pg$se
  pg$ci_hi <- pg$predicted + 1.96 * pg$se

  obs <- fdc %>%
    mutate(vy_r = round(visit_year)) %>%
    group_by(Cohort, vy_r) %>%
    summarise(mn = mean(score, na.rm = TRUE),
              se = sd(score, na.rm = TRUE) / sqrt(n()), n = n(), .groups = "drop") %>%
    filter(n >= 5)
  
 n_tbl <- fdc %>%
    mutate(vy_r = round(visit_year)) %>%
    group_by(Cohort, time = vy_r) %>%
    summarise(n = n_distinct(participant_id), .groups = "drop")
    n_tbl$n[n_tbl$time == 0 & n_tbl$Cohort == "MC1R PD"] <- 231
    n_tbl$n[n_tbl$time == 0 & n_tbl$Cohort == "SPD"] <- 152

  cols <- c("SPD" = "#2166AC", "MC1R PD" = "#B2182B")

  int_nm_fig <- grep("visit_year:Cohort|Cohort.*:visit_year", names(fe), value = TRUE)[1]
  ref_slope_fig <- fe["visit_year"]
  mc1r_slope_fig <- ref_slope_fig + fe[int_nm_fig]
  bd_fig <- fe[int_nm_fig]
  bd_se_fig <- sqrt(diag(vm))[int_nm_fig]
  pv_fig <- summary(fig_mod)$tTable[int_nm_fig, "p-value"]
  pct_fig <- round(100 * bd_fig / abs(ref_slope_fig), 1)
  n_spd_fig <- 152
  n_mc1r_fig <- 231

  stat_label <- sprintf(
      "Non-carriers: %0.2f pts/yr (n=%d)<br><i>MC1R</i> carriers: %0.2f pts/yr (n=%d)<br>\u03b2=%0.2f (%0.2f, %0.2f); %+0.1f%% faster; p=%s",
      ref_slope_fig, n_spd_fig, mc1r_slope_fig, n_mc1r_fig,
      bd_fig, bd_fig - 1.96*bd_se_fig, bd_fig + 1.96*bd_se_fig,
      pct_fig, fmt_p(pv_fig))

  p_main <- ggplot() +
    geom_ribbon(data = pg, aes(visit_year, ymin = ci_lo, ymax = ci_hi, fill = Cohort), alpha = 0.2) +
    geom_point(data = obs, aes(vy_r, mn, color = Cohort), size = 3) +
    geom_errorbar(data = obs, aes(vy_r, mn, ymin = mn - 1.96*se, ymax = mn + 1.96*se, color = Cohort),
                  width = 0.2, linewidth = 0.8) +
    geom_line(data = pg, aes(visit_year, predicted, color = Cohort), linewidth = 1.5) +
    annotate("richtext", x = 0.05, y = max(pg$ci_hi, na.rm = TRUE) * 1.08,
             label = stat_label, hjust = 0, vjust = 1,
             size = 3.5, fill = "white", color = "black",
             label.size = 0.5, label.r = unit(0.15, "lines"),
             family = "sans") +
    scale_color_manual(values = cols, labels = c("SPD" = "Non-carriers", "MC1R PD" = "<i>MC1R</i> carriers")) +
    scale_fill_manual(values = cols, labels = c("SPD" = "Non-carriers", "MC1R PD" = "<i>MC1R</i> carriers")) +
    scale_x_continuous(breaks = seq(0, 20, 1), limits = c(0, 12.1), expand = expansion(mult = c(0.02, 0.05))) +
    scale_y_continuous(breaks = seq(0, 100, 5), expand = expansion(mult = c(0.02, 0.05))) +
    labs(x = "Years since Baseline", y = "MDS-UPDRS III Score", color = "Group",
        title = "PPMI Cohort") +
    theme_classic(base_size = 14) +
    theme(legend.position = c(0.85, 0.15),
          legend.background = element_rect(fill = "white", color = "gray80"),
          legend.title = element_blank(),
          legend.text = element_markdown(),
          axis.title = element_text(face = "bold"),
          axis.text = element_text(color = "black"),
          axis.title.y = element_text(face = "bold", vjust = -14),
          plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
          panel.grid.major = element_line(color = "gray90", linewidth = 0.3),
          plot.margin = margin(5, 10, -5, 10)) +
    guides(fill = "none") +
    coord_cartesian(ylim = c(18, 60))
  
  n_tbl$Cohort <- factor(n_tbl$Cohort, levels = c("MC1R PD", "SPD"))
  p_n <- ggplot(n_tbl, aes(x = time, y = Cohort, label = n)) +
    geom_text(aes(color = Cohort), size = 3.5, fontface = "plain") +
    scale_color_manual(values = cols, guide = "none") +
    scale_x_continuous(breaks = seq(0, 12, 1), limits = c(0, 12.1),
                       expand = expansion(mult = c(0.02, 0.05))) +
    scale_y_discrete(labels = c("MC1R PD" = expression(paste(italic("MC1R"), " carriers")),
                                "SPD" = "Non-carriers")) +
    labs(x = NULL, y = NULL) +
    theme_void(base_size = 11) +
    theme(axis.text.y = element_text(hjust = 1, face = "bold", color = "gray30",
                                     margin = margin(r = 10)),
          axis.text.x = element_blank(),
          plot.title = element_text(size = 10, face = "bold", hjust = -0.15,
                                    color = "gray30", margin = margin(b = 3)),
          plot.margin = margin(-5, 10, 5, 60)) +
    ggtitle("No. of participants")

    fig <- p_main / p_n + plot_layout(heights = c(8, 1))
  } else {
    fig <- p_main
  }

  print(fig)
  ggsave(file.path(config$out_dir, "Figure_UPDRS3_Progression.png"), fig, width = 8, height = 6, dpi = 600)
  ggsave(file.path(config$out_dir, "Figure_UPDRS3_Progression.pdf"), fig, width = 8, height = 6, device = cairo_pdf)
```
```{r}
fdc_8 <- fdc %>% filter(visit_year <= 8.5)

fig_mod_8 <- tryCatch({
  lme(fig_fml, data = fdc_8, random = ~ visit_year + 1 | participant_id,
      weights = varIdent(form = ~ 1 | Cohort),
      control = lmeControl(opt = "optim", maxIter = 200, msMaxIter = 200), na.action = na.omit)
}, error = function(e) NULL)

if (!is.null(fig_mod_8)) {
  fe8 <- fixef(fig_mod_8)
  vm8 <- vcov(fig_mod_8)
  yrs8 <- seq(0, 8, by = 0.5)

  cov_val8 <- fe8["baseline_score"] * mean(fdc_8$baseline_score, na.rm = TRUE) +
    fe8["Age_onset"] * mean(fdc_8$Age_onset, na.rm = TRUE) +
    fe8["LEDD"] * mean(fdc_8$LEDD, na.rm = TRUE)

  int_nm8 <- grep("visit_year:Cohort|Cohort.*:visit_year", names(fe8), value = TRUE)[1]

  pg8 <- data.frame(
    visit_year = rep(yrs8, 2),
    Cohort = factor(rep(c("SPD", "MC1R PD"), each = length(yrs8)), levels = c("SPD", "MC1R PD")))
  pg8$predicted <- ifelse(
    pg8$Cohort == "SPD",
    fe8["(Intercept)"] + fe8["visit_year"] * pg8$visit_year + cov_val8,
    fe8["(Intercept)"] + fe8["CohortMC1R PD"] +
      (fe8["visit_year"] + fe8[int_nm8]) * pg8$visit_year + cov_val8)

  pg8$se <- NA
  for (i in 1:nrow(pg8)) {
    t <- pg8$visit_year[i]
    if (pg8$Cohort[i] == "SPD") {
      vp <- vm8["(Intercept)","(Intercept)"] + t^2*vm8["visit_year","visit_year"] +
        2*t*vm8["(Intercept)","visit_year"]
    } else {
      vp <- vm8["(Intercept)","(Intercept)"] + vm8["CohortMC1R PD","CohortMC1R PD"] +
        t^2*vm8["visit_year","visit_year"] + t^2*vm8[int_nm8, int_nm8] +
        2*vm8["(Intercept)","CohortMC1R PD"] + 2*t*vm8["(Intercept)","visit_year"] +
        2*t*vm8["(Intercept)", int_nm8] + 2*t*vm8["CohortMC1R PD","visit_year"] +
        2*t*vm8["CohortMC1R PD", int_nm8] + 2*t^2*vm8["visit_year", int_nm8]
    }
    pg8$se[i] <- sqrt(max(vp, 0))
  }
  pg8$ci_lo <- pg8$predicted - 1.96 * pg8$se
  pg8$ci_hi <- pg8$predicted + 1.96 * pg8$se

  obs8 <- fdc_8 %>%
    mutate(vy_r = round(visit_year)) %>%
    group_by(Cohort, vy_r) %>%
    summarise(mn = mean(score, na.rm = TRUE),
              se = sd(score, na.rm = TRUE) / sqrt(n()), n = n(), .groups = "drop") %>%
    filter(n >= 5)

  n_tbl8 <- fdc_8 %>%
    mutate(vy_r = round(visit_year)) %>%
    group_by(Cohort, time = vy_r) %>%
    summarise(n = n_distinct(participant_id), .groups = "drop")

  ref8 <- fe8["visit_year"]
  mc1r8 <- ref8 + fe8[int_nm8]
  bd8 <- fe8[int_nm8]
  bd_se8 <- sqrt(diag(vm8))[int_nm8]
  pv8 <- summary(fig_mod_8)$tTable[int_nm8, "p-value"]
  pct8 <- round(100 * bd8 / abs(ref8), 1)
  n_spd8 <- 152
  n_mc1r8 <- 231

  stat_label8 <- sprintf(
    "Non-carriers: %0.2f pts/yr (n=%d)<br><i>MC1R</i> carriers: %0.2f pts/yr (n=%d)<br>\u03b2=%0.2f (%0.2f, %0.2f); %+.1f%% faster; p=%s",
    ref8, n_spd8, mc1r8, n_mc1r8,
    bd8, bd8 - 1.96*bd_se8, bd8 + 1.96*bd_se8,
    pct8, fmt_p(pv8))

  cat(sprintf("8-year sensitivity: beta=%.3f, p=%.4f, %+.1f%% faster\n", bd8, pv8, pct8))

  cols <- c("SPD" = "#2166AC", "MC1R PD" = "#B2182B")

  p_main8 <- ggplot() +
    geom_ribbon(data = pg8, aes(visit_year, ymin = ci_lo, ymax = ci_hi, fill = Cohort), alpha = 0.2) +
    geom_point(data = obs8, aes(vy_r, mn, color = Cohort), size = 3) +
    geom_errorbar(data = obs8, aes(vy_r, mn, ymin = mn - 1.96*se, ymax = mn + 1.96*se, color = Cohort),
                  width = 0.2, linewidth = 0.8) +
    geom_line(data = pg8, aes(visit_year, predicted, color = Cohort), linewidth = 1.5) +
    annotate("richtext", x = 0.05, y = 45,
             label = stat_label8, hjust = 0, vjust = 1,
             size = 3.5, fill = "white", color = "black",
             label.size = 0.5, label.r = unit(0.15, "lines"),
             family = "sans") +
    scale_color_manual(values = cols, labels = c("SPD" = "Non-carriers", "MC1R PD" = "<i>MC1R</i> carriers")) +
    scale_fill_manual(values = cols, labels = c("SPD" = "Non-carriers", "MC1R PD" = "<i>MC1R</i> carriers")) +
    scale_x_continuous(breaks = seq(0, 8, 1), limits = c(0, 8.1), expand = expansion(mult = c(0.02, 0.05))) +
    scale_y_continuous(breaks = seq(0, 45, 5), expand = expansion(mult = c(0.02, 0.05))) +
    labs(x = "Years since Baseline", y = "MDS-UPDRS III Score",
         title = "PPMI Cohort (Censored at 8 Years)") +
    theme_classic(base_size = 14) +
    theme(legend.position = c(0.85, 0.15),
          legend.background = element_rect(fill = "white", color = "gray80"),
          legend.title = element_blank(),
          legend.text = element_markdown(),
          axis.title.x = element_text(face = "bold"),
          axis.title.y = element_text(face = "bold", vjust = -14),
          axis.text = element_text(color = "black"),
          plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
          panel.grid.major = element_line(color = "gray90", linewidth = 0.3),
          plot.margin = margin(5, 10, -5, 10)) +
    guides(fill = "none") +
    coord_cartesian(ylim = c(18, 45))

  n_tbl8$Cohort <- factor(n_tbl8$Cohort, levels = c("MC1R PD", "SPD"))
    n_tbl8$n[n_tbl8$time == 0 & n_tbl8$Cohort == "MC1R PD"] <- 231
    n_tbl8$n[n_tbl8$time == 0 & n_tbl8$Cohort == "SPD"] <- 152
  p_n8 <- ggplot(n_tbl8, aes(x = time, y = Cohort, label = n)) +
    geom_text(aes(color = Cohort), size = 3.5, fontface = "plain") +
    scale_color_manual(values = cols, guide = "none") +
    scale_x_continuous(breaks = seq(0, 8, 1), limits = c(0, 8.1),
                       expand = expansion(mult = c(0.02, 0.05))) +
    scale_y_discrete(labels = c("MC1R PD" = expression(paste(italic("MC1R"), " carriers")),
                                "SPD" = "Non-carriers")) +
    labs(x = NULL, y = NULL) +
    theme_void(base_size = 11) +
    theme(axis.text.y = element_text(hjust = 1, face = "bold", color = "gray30",
                                     margin = margin(r = 10)),
          axis.text.x = element_blank(),
          plot.title = element_text(size = 10, face = "bold", hjust = -0.15,
                                    color = "gray30", margin = margin(b = 3)),
          plot.margin = margin(-5, 10, 5, 60)) +
    ggtitle("No. of participants")

  fig_8yr <- p_main8 / p_n8 + plot_layout(heights = c(8, 1))

  print(fig_8yr)
  ggsave(file.path(config$out_dir, "Figure_UPDRS3_8yr_Sensitivity.png"), fig_8yr, width = 8, height = 6, dpi = 600)
  ggsave(file.path(config$out_dir, "Figure_UPDRS3_8yr_Sensitivity.pdf"), fig_8yr, width = 8, height = 6, device = cairo_pdf)
}
```

## Supplemental: Person-Level Slope Distributions

```{r slope-density, fig.width=8, fig.height=5}
if (!is.null(fig_mod)) {
  re <- ranef(fig_mod)
  fe <- fixef(fig_mod)
  int_nm <- grep("visit_year:Cohort|Cohort.*:visit_year", names(fe), value = TRUE)[1]

  slope_df <- data.frame(
    participant_id = rownames(re),
    random_slope = re$visit_year
  ) %>%
    left_join(fdc %>% distinct(participant_id, Cohort), by = "participant_id") %>%
    filter(!is.na(Cohort)) %>%
    mutate(person_slope = ifelse(
      Cohort == "SPD",
      fe["visit_year"] + random_slope,
      fe["visit_year"] + fe[int_nm] + random_slope
    ))

  cols <- c("SPD" = "#2166AC", "MC1R PD" = "#B2182B")

  p_dens <- ggplot(slope_df, aes(x = person_slope, fill = Cohort, color = Cohort)) +
    geom_density(alpha = 0.3, linewidth = 0.8) +
    geom_vline(data = slope_df %>% group_by(Cohort) %>% summarise(mn = mean(person_slope)),
               aes(xintercept = mn, color = Cohort), linewidth = 1, linetype = "dashed",
               show.legend = FALSE) +
    scale_fill_manual(values = cols, labels = c("SPD" = "Non-carriers", "MC1R PD" = "<i>MC1R</i> carriers")) +
    scale_color_manual(values = cols, labels = c("SPD" = "Non-carriers", "MC1R PD" = "<i>MC1R</i> carriers")) +
    labs(x = "Estimated Person-Level Slope (MDS-UPDRS III pts/yr)",
         y = "Density",
         title = "PPMI: Distribution of Individual Motor Progression Rates") +
    theme_classic(base_size = 14) +
    theme(legend.position = c(0.85, 0.85),
          legend.background = element_rect(fill = "white", color = "gray80"),
          legend.title = element_blank(),
          legend.text = element_markdown(),
          axis.title = element_text(face = "bold"),
          axis.text = element_text(color = "black"),
          plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
          panel.grid.major = element_line(color = "gray90", linewidth = 0.3)) +
    guides(fill = guide_legend(override.aes = list(color = NA, linetype = 0)),
           color = "none")

  print(p_dens)
  ggsave(file.path(config$out_dir, "Supplemental_Slope_Density_PPMI.png"), p_dens, width = 8, height = 5, dpi = 600)
  ggsave(file.path(config$out_dir, "Supplemental_Slope_Density_PPMI.pdf"), p_dens, width = 8, height = 5, device = cairo_pdf)
}
```

```{r slope-vs-treatment, fig.width=8, fig.height=5}
if (!is.null(fig_mod) && exists("slope_df")) {
  max_fu <- fdc %>%
    group_by(participant_id) %>%
    summarise(max_year = max(visit_year, na.rm = TRUE), .groups = "drop")

  trt_time <- d_u3 %>%
    filter(participant_id %in% slope_df$participant_id, !is.na(LEDD), LEDD > 0) %>%
    group_by(participant_id) %>%
    slice_min(visit_year, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    select(participant_id, trt_year = visit_year)

  slope_trt <- slope_df %>%
    left_join(max_fu, by = "participant_id") %>%
    left_join(trt_time, by = "participant_id") %>%
    mutate(
      started_trt = !is.na(trt_year),
      time_to_trt = coalesce(trt_year, max_year)
    ) %>%
    filter(is.finite(time_to_trt))

  slope_trt <- slope_trt %>%
    mutate(group_label = interaction(
      ifelse(Cohort == "MC1R PD", "<i>MC1R</i> carriers", "Non-carriers"),
      ifelse(started_trt, "Started treatment", "Not yet treated"),
      sep = " - "
    ))

  grp_cols <- c("<i>MC1R</i> carriers - Started treatment" = "#B2182B",
                "<i>MC1R</i> carriers - Not yet treated" = "#B2182B",
                "Non-carriers - Started treatment" = "#2166AC",
                "Non-carriers - Not yet treated" = "#2166AC")
  grp_shapes <- c("<i>MC1R</i> carriers - Started treatment" = 16,
                   "<i>MC1R</i> carriers - Not yet treated" = 1,
                   "Non-carriers - Started treatment" = 16,
                   "Non-carriers - Not yet treated" = 1)
  grp_order <- c("<i>MC1R</i> carriers - Started treatment", "<i>MC1R</i> carriers - Not yet treated",
                  "Non-carriers - Started treatment", "Non-carriers - Not yet treated")
  slope_trt$group_label <- factor(slope_trt$group_label, levels = grp_order)

  p_trt <- ggplot(slope_trt, aes(x = time_to_trt, y = person_slope,
                                  color = group_label, shape = group_label)) +
    geom_point(alpha = 0.5, size = 2) +
    geom_smooth(data = slope_trt, aes(x = time_to_trt, y = person_slope, color = Cohort),
                method = "lm", se = FALSE, linewidth = 0.8, linetype = "dashed",
                show.legend = FALSE, inherit.aes = FALSE) +
    scale_color_manual(values = c(grp_cols, "SPD" = "#2166AC", "MC1R PD" = "#B2182B"),
                       breaks = grp_order, name = NULL) +
    scale_shape_manual(values = grp_shapes, breaks = grp_order, name = NULL) +
    labs(x = "Time to Treatment Initiation (years)",
         y = "Estimated Person-Level Slope (pts/yr)",
         title = "PPMI: Progression Rate vs. Time to Treatment Initiation") +
    theme_classic(base_size = 14) +
    theme(legend.position = "right",
          legend.background = element_rect(fill = "white", color = "gray80"),
          legend.text = element_markdown(),
          axis.title = element_text(face = "bold"),
          axis.text = element_text(color = "black"),
          plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
          panel.grid.major = element_line(color = "gray90", linewidth = 0.3)) +
    coord_cartesian(xlim = c(0, 3))

  print(p_trt)
  ggsave(file.path(config$out_dir, "Supplemental_Slope_vs_Treatment_PPMI.png"), p_trt, width = 8, height = 5, dpi = 600)
  ggsave(file.path(config$out_dir, "Supplemental_Slope_vs_Treatment_PPMI.pdf"), p_trt, width = 8, height = 5, device = cairo_pdf)
}
```

```{r attrition, fig.width=8, fig.height=5}
att <- fdc %>%
  mutate(yr = floor(visit_year)) %>%
  group_by(Cohort, yr) %>%
  summarise(n = n_distinct(participant_id), .groups = "drop")

cols <- c("SPD" = "#2166AC", "MC1R PD" = "#B2182B")

p_att <- ggplot(att, aes(x = yr, y = n, color = Cohort, group = Cohort)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  geom_text(aes(label = n), vjust = -1, size = 3, show.legend = FALSE) +
  scale_color_manual(values = cols, labels = c("SPD" = "Non-carriers", "MC1R PD" = "<i>MC1R</i> carriers")) +
  scale_x_continuous(breaks = seq(0, 12, 1)) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
  labs(x = "Year of Follow-up", y = "Participants with Assessments",
       title = "PPMI: Sample Size by Year of Follow-up") +
  theme_classic(base_size = 14) +
  theme(legend.position = c(0.85, 0.85),
        legend.background = element_rect(fill = "white", color = "gray80"),
        legend.title = element_blank(),
        legend.text = element_markdown(),
        axis.title = element_text(face = "bold"),
        axis.text = element_text(color = "black"),
        plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
        panel.grid.major = element_line(color = "gray90", linewidth = 0.3))

print(p_att)
ggsave(file.path(config$out_dir, "Supplemental_Attrition_PPMI.png"), p_att, width = 8, height = 5, dpi = 600)
ggsave(file.path(config$out_dir, "Supplemental_Attrition_PPMI.pdf"), p_att, width = 8, height = 5, device = cairo_pdf)
```

## Demographics Table 1

```{r table1}
u3_summ <- d_u3 %>%
  group_by(participant_id, Cohort) %>%
  summarise(bl_updrs3 = first(score, order_by = visit_year),
            max_fu = max(visit_year, na.rm = TRUE), .groups = "drop")
moca_summ <- d_moca %>%
  group_by(participant_id, Cohort) %>%
  summarise(bl_moca = first(score, order_by = visit_year), .groups = "drop")
ledd_summ <- d_u3 %>%
  group_by(participant_id, Cohort) %>%
  summarise(bl_ledd = first(LEDD, order_by = visit_year), .groups = "drop") %>%
  mutate(on_dopa = !is.na(bl_ledd) & bl_ledd > 0)

dem_tbl <- dem %>%
  left_join(u3_summ, by = c("Subject_ID" = "participant_id", "Cohort")) %>%
  left_join(moca_summ, by = c("Subject_ID" = "participant_id", "Cohort")) %>%
  left_join(ledd_summ, by = c("Subject_ID" = "participant_id", "Cohort")) %>%
  mutate(
    Race_detail = case_when(
      Race == "White" ~ "White",
      grepl("Asian", Race, ignore.case = TRUE) ~ "Asian",
      grepl("Black", Race, ignore.case = TRUE) ~ "Black",
      grepl("Multi", Race, ignore.case = TRUE) ~ "Multiracial",
      TRUE ~ "Unknown"),
    Hispanic     = grepl("Hispanic or Latino", Ethnicity, ignore.case = TRUE),
    Parent_PD    = grepl("^Yes", `Parent with PD`, ignore.case = TRUE),
    Other_Rel_PD = grepl("^Yes", `Other Relative with PD`, ignore.case = TRUE),
    Melanoma     = grepl("Melanoma", `History of Skin Cancer`, ignore.case = TRUE),
    SAA_pos      = `SAA Result` == "Positive",
    SAA_neg      = `SAA Result` == "Negative",
    SAA_unavail  = !(`SAA Result` %in% c("Positive", "Negative")),
    is_hom       = Homozygote == "Yes"
  )

build_col <- function(d) {
  n <- nrow(d)
  pct <- function(x) sprintf("%d (%.1f%%)", sum(x, na.rm = TRUE), 100 * mean(x, na.rm = TRUE))
  med_iqr <- function(x) {
    x <- x[!is.na(x)]; if (!length(x)) return("\u2014")
    sprintf("%.1f (%.1f\u2013%.1f)", median(x), quantile(x, .25), quantile(x, .75))
  }
  mn_sd <- function(x) {
    x <- x[!is.na(x)]; if (!length(x)) return("\u2014")
    sprintf("%.1f \u00b1 %.1f", mean(x), sd(x))
  }
  saa_pos_n <- sum(d$SAA_pos, na.rm = TRUE)
  saa_neg_n <- sum(d$SAA_neg, na.rm = TRUE)
  saa_known <- saa_pos_n + saa_neg_n

  c(N = as.character(n),
    Homozygotes = pct(d$is_hom),
    `Age at baseline (years)` = mn_sd(d$Age_baseline),
    `Years since diagnosis` = mn_sd(d$Diagnosis),
    `Age at PD onset (years)` = mn_sd(d$`Age at PD onset`),
    Male = pct(d$Sex == "Male"),
    Race = "",
    `  White` = pct(d$Race_detail == "White"),
    `  Asian` = pct(d$Race_detail == "Asian"),
    `  Black` = pct(d$Race_detail == "Black"),
    `  Multiracial` = pct(d$Race_detail == "Multiracial"),
    `  Unknown` = pct(d$Race_detail == "Unknown"),
    `Hispanic or Latino` = pct(d$Hispanic),
    `Relatives with PD` = "",
    `  Parent` = pct(d$Parent_PD),
    `  Other` = pct(d$Other_Rel_PD),
    `SAA status` = "",
    `  Positive` = pct(d$SAA_pos),
    `  Negative` = pct(d$SAA_neg),
    `  Unavailable` = pct(d$SAA_unavail),
    `SAA positivity` = if (saa_known > 0)
      sprintf("%d/%d (%.1f%%)", saa_pos_n, saa_known, 100*saa_pos_n/saa_known) else "\u2014",
    `History of melanoma` = pct(d$Melanoma),
    `MDS-UPDRS III at baseline` = mn_sd(d$bl_updrs3),
    `MoCA at baseline` = mn_sd(d$bl_moca),
    `On dopaminergic medication` = pct(d$on_dopa),
    `Average follow-up (years)` = mn_sd(d$max_fu))
}

cohorts <- list(
  SPD = dem_tbl %>% filter(Cohort == "SPD"),
  `MC1R PD` = dem_tbl %>% filter(Cohort == "MC1R PD"),
  `LRRK2 PD` = dem_tbl %>% filter(Cohort == "LRRK2 PD"),
  `MC1R + LRRK2 PD` = dem_tbl %>% filter(Cohort == "MC1R + LRRK2 PD"),
  `GBA PD` = dem_tbl %>% filter(Cohort == "GBA PD"),
  `MC1R + GBA PD` = dem_tbl %>% filter(Cohort == "MC1R + GBA PD"),
  All = dem_tbl)

table1 <- data.frame(Characteristic = names(build_col(dem_tbl)))
for (cn in names(cohorts)) table1[[cn]] <- build_col(cohorts[[cn]])
print(knitr::kable(table1, caption = "Table 1"))
```

## Output

```{r output}
fmt_out <- function(df) {
  df %>%
    mutate(across(where(is.numeric) & matches("Slope|Beta|SE|CI|Mean|Diff"), ~ round(., 3)),
           across(where(is.numeric) & matches("P_value"), ~ round(., 4)))
}

all_bio <- bind_rows(saa_bl_df, csf_bl_df, asyn_bl, nfl_df, gfap_df)
sec_all <- bind_rows(u1_res, moca_res, dat_c_res, dat_p_res, u2_res, u4_res, comb_res, total_res)

wb <- createWorkbook()
addWorksheet(wb, "Table1"); writeData(wb, "Table1", table1)
addWorksheet(wb, "UPDRS3"); writeData(wb, "UPDRS3", fmt_out(u3_df))
addWorksheet(wb, "UPDRS3_ON"); writeData(wb, "UPDRS3_ON", fmt_out(u3on_df))
addWorksheet(wb, "UPDRS3_Stratified"); writeData(wb, "UPDRS3_Stratified", fmt_out(u3_strat_df))
addWorksheet(wb, "UPDRS3_Subscores"); writeData(wb, "UPDRS3_Subscores", sub_off_df)
addWorksheet(wb, "UPDRS3_Subscores_ON"); writeData(wb, "UPDRS3_Subscores_ON", sub_on_df)
addWorksheet(wb, "UPDRS3_Subscores_All"); writeData(wb, "UPDRS3_Subscores_All", sub_all_df)

addWorksheet(wb, "Cocarrier_Full")
writeData(wb, "Cocarrier_Full", bind_rows(cc_off_full, cc_on_full) %>% fmt_out())
addWorksheet(wb, "Cocarrier_Penetrance")
writeData(wb, "Cocarrier_Penetrance", bind_rows(cc_pen_off, cc_pen_on) %>% fmt_out())

outcome_sheets <- c(
  "MDS-UPDRS I" = "UPDRS1", "MDS-UPDRS II" = "UPDRS2", "MDS-UPDRS IV" = "UPDRS4",
  "MoCA" = "MoCA", "DAT-SPECT Caudate" = "DAT_Caudate", "DAT-SPECT Putamen" = "DAT_Putamen",
  "Combined MDS-UPDRS I-III" = "UPDRS_Combined", "Total MDS-UPDRS I-IV" = "UPDRS_Total")
for (ol in names(outcome_sheets)) {
  sn <- outcome_sheets[ol]
  rows <- sec_all %>% filter(Outcome == ol)
  if (nrow(rows) > 0) { addWorksheet(wb, sn); writeData(wb, sn, fmt_out(rows)) }
}

addWorksheet(wb, "CSF_Biomarkers"); writeData(wb, "CSF_Biomarkers", fmt_out(all_bio))
if (nrow(saa_pen) > 0) { addWorksheet(wb, "SAA_Penetrance"); writeData(wb, "SAA_Penetrance", fmt_out(saa_pen)) }
if (AGE_INT_OK) { addWorksheet(wb, "Age_Interaction"); writeData(wb, "Age_Interaction", age_int) }
if (SEX_INT_OK) { addWorksheet(wb, "Sex_Interaction"); writeData(wb, "Sex_Interaction", sex_int) }
if (DUR_INT_OK) { addWorksheet(wb, "Duration_Interaction"); writeData(wb, "Duration_Interaction", dur_int) }

if (GEN_IND_OK) {
  addWorksheet(wb, "Genetic_Ind_Coefs"); writeData(wb, "Genetic_Ind_Coefs", gi_coef)
  if (nrow(gi_strata) > 0) { addWorksheet(wb, "Genetic_Ind_Strata"); writeData(wb, "Genetic_Ind_Strata", gi_strata) }
  if (gi_m3a$success) {
    m3a_tbl <- as.data.frame(summary(gi_m3a$model)$tTable)
    m3a_tbl$Term <- rownames(m3a_tbl)
    m3a_tbl <- m3a_tbl[, c("Term", setdiff(names(m3a_tbl), "Term"))]
    addWorksheet(wb, "Genetic_Ind_M3a"); writeData(wb, "Genetic_Ind_M3a", m3a_tbl)
  }
  if (gi_m3b$success) {
    m3b_tbl <- as.data.frame(summary(gi_m3b$model)$tTable)
    m3b_tbl$Term <- rownames(m3b_tbl)
    m3b_tbl <- m3b_tbl[, c("Term", setdiff(names(m3b_tbl), "Term"))]
    addWorksheet(wb, "Genetic_Ind_M3b"); writeData(wb, "Genetic_Ind_M3b", m3b_tbl)
  }
}

if (nrow(scopa_df) > 0) { addWorksheet(wb, "SCOPA_AUT"); writeData(wb, "SCOPA_AUT", fmt_out(scopa_df)) }
if (nrow(gds_df) > 0) { addWorksheet(wb, "GDS"); writeData(wb, "GDS", fmt_out(gds_df)) }
if (nrow(subtype_df) > 0) { addWorksheet(wb, "TD_PIGD"); writeData(wb, "TD_PIGD", fmt_out(subtype_df)) }
if (nrow(trend_results) > 0) { addWorksheet(wb, "Dose_Penetrance_Trends"); writeData(wb, "Dose_Penetrance_Trends", trend_results) }
if (nrow(fp_df) > 0) {
  addWorksheet(wb, "Fast_Progressor"); writeData(wb, "Fast_Progressor", fp_df)
  if (nrow(fp_trend) > 0) writeData(wb, "Fast_Progressor", fp_trend, startRow = nrow(fp_df) + 3)
}
if (nrow(ess_df) > 0) { addWorksheet(wb, "ESS"); writeData(wb, "ESS", fmt_out(ess_df)) }

if (length(sens_results) > 0) {
  addWorksheet(wb, "Sensitivity"); writeData(wb, "Sensitivity", bind_rows(sens_results))
}

saveWorkbook(wb, file.path(config$out_dir, paste0("MC1R_Results_", Sys.Date(), ".xlsx")), overwrite = TRUE)
```

```{r summary}
table(dem$Cohort)
pm <- u3_df %>% filter(Analysis_type == "primary", LEDD_adjusted == "Yes", Group == "MC1R PD")
ps <- u3_df %>% filter(Analysis_type == "primary", LEDD_adjusted == "Yes", Group == "SPD")
if (nrow(pm) > 0 && nrow(ps) > 0)
  cat(sprintf("SPD: %.2f pts/yr | MC1R: %.2f pts/yr | diff=%.2f (%.1f%%), p=%s\n",
              ps$Slope, pm$Slope, pm$Beta_diff, pm$Pct_diff, fmt_p(pm$P_value)))
```

```{r session}
sessionInfo()
```
